# Networks of Change: Systems Thinking in Connected Worlds

The previous post treated uncertainty as something that perturbs a single evolving state variable. That is a useful starting point, but it omits a feature of most biological, psychological, and social systems: variables rarely evolve alone. Neurons influence other neurons, symptoms alter one another’s trajectories, infections follow patterns of contact, and beliefs or behaviours spread through social ties. In each case, the dynamics of a unit depend partly on its own state and partly on the states of its neighbours.

Network dynamics provides a language for this situation. Rather than asking only whether one variable is stable, noisy, or close to a bifurcation, it asks how a collection of coupled variables produces collective behaviour. Synchronization, cascades, clustering, and abrupt global transitions can then emerge from interactions even when no individual component has those properties in isolation. Complex-network research formalizes these questions using network structure, including degree distributions, clustering, and correlations, alongside the dynamics occurring on the network. [arxiv](https://arxiv.org/abs/cond-mat/0303516)

## From a State to a Network

Let a network contain $N$ nodes, indexed by $i=1,\ldots,N$. Each node has a state $x_i(t)$, which might represent neural activity, symptom severity, infection probability, opinion, or another quantity whose meaning depends on the application. The adjacency matrix $A$ records which nodes are connected:

$$
A_{ij} =
\begin{cases}
1, & \text{if nodes } i \text{ and } j \text{ are connected},\\
0, & \text{otherwise}.
\end{cases}
$$

For weighted networks, $A_{ij}$ need not be binary: it can encode contact frequency, connection strength, similarity, or estimated functional connectivity. In directed networks, $A_{ij}$ and $A_{ji}$ may differ, so influence has an orientation.

A broad class of coupled dynamical systems can be written as

$$
\frac{dx_i}{dt}
=
f(x_i)
+
\kappa \sum_{j=1}^{N} A_{ij} g(x_j-x_i),
$$

where $f(x_i)$ describes the intrinsic dynamics of node $i$, $g$ specifies how differences between connected nodes are translated into influence, and $\kappa$ is a global coupling strength. The equation is deliberately generic. Its value is that it separates two distinct questions: what each unit would do alone, and what the topology of its relationships makes possible.

If $g(z)=z$, coupling is diffusive:

$$
\frac{dx_i}{dt}
=
f(x_i)
+
\kappa \sum_{j=1}^{N} A_{ij}(x_j-x_i).
$$

A node is pulled toward the states of its connected neighbours. This is not necessarily a literal mechanism of imitation; it is often a coarse model of exchange, shared input, regulation, or mutual constraint.

## Graph Structure Is Dynamical Structure

The adjacency matrix is more than a storage device for edges. It determines which signals can propagate, which regions can become locally coherent, and which nodes can exert disproportionate influence.

The degree of node $i$ in an undirected binary network is

$$
k_i = \sum_{j=1}^{N} A_{ij}.
$$

Nodes with large $k_i$ are often called hubs. A hub can have an outsized role in diffusion or synchronization, although degree alone is not enough to determine influence: edge weights, directionality, timing, and the nonlinear form of coupling also matter. Network science has shown that properties such as small-world organization, clustering, degree heterogeneity, and assortative mixing can substantially alter the dynamical processes that run over an otherwise similar number of nodes and edges. [arxiv](https://arxiv.org/abs/cond-mat/0303516)

It is often useful to define the degree matrix $D$, where $D_{ii}=k_i$, and the graph Laplacian

$$
L = D-A.
$$

For diffusive coupling, the network term can be written compactly as

$$
\dot{\mathbf{x}} = \mathbf{f}(\mathbf{x}) - \kappa L\mathbf{x}.
$$

The Laplacian makes a basic point visible: coupling penalizes differences across edges. If all connected nodes have identical values, then $L\mathbf{x}=\mathbf{0}$, and the diffusive term vanishes. Whether that uniform state is stable depends on the node dynamics, coupling, and the spectrum of $L$, not on topology alone.

## Synchronization as a Collective Transition

Synchronization occurs when interacting oscillatory units develop a common rhythm, or at least a stable relationship among their phases. It is familiar in examples ranging from coordinated neuronal oscillations to power grids and circadian processes, but the word should not be treated as automatically positive. In some systems, synchrony supports coherent function; in others, excessive synchronization may be pathological or reduce flexibility.

A standard model is the networked Kuramoto system:

$$
\frac{d\theta_i}{dt}
=
\omega_i
+
K \sum_{j=1}^{N} A_{ij}\sin(\theta_j-\theta_i).
$$

Here $\theta_i$ is the phase of oscillator $i$, $\omega_i$ is its intrinsic frequency, and $K$ scales the coupling. When $K$ is small, differences in $\omega_i$ dominate and phases drift apart. As $K$ increases, some oscillators can phase-lock; under suitable conditions, a large fraction of the network may become coherent. Synchronization in networks is a fundamental phenomenon, but its onset and transient development depend strongly on network structure rather than coupling strength alone. [piccardi.faculty.polimi](https://piccardi.faculty.polimi.it/VarieCsr/Papers/Arenas2008.pdf)

A common measure of global phase coherence is the Kuramoto order parameter:

$$
R(t)e^{i\Psi(t)}
=
\frac{1}{N}
\sum_{j=1}^{N}e^{i\theta_j(t)}.
$$

The scalar $R(t)$ lies between 0 and 1. Values near 0 indicate dispersed phases; values near 1 indicate near-perfect alignment. It is a summary statistic, though, and may conceal important structure. Two communities can each be internally synchronized while remaining out of phase with one another, producing an intermediate global $R$ that does not describe either local organization very well.

## R Simulation: Network Structure

The first code block builds three small networks with the same number of nodes but different topology: a ring, a random graph, and a modular graph. The point is not that any one resembles a real brain or social network. It is to make topology concrete before dynamics are placed on top of it.

``` {r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(igraph)
library(patchwork)
library(viridis)

set.seed(42)

n <- 30

# 1. Ring lattice: each node connects to its two nearest neighbours
g_ring <- make_ring(n, directed = FALSE)

# 2. Erdos-Renyi random graph with roughly comparable mean degree
g_random <- sample_gnp(n, p = 2 / (n - 1), directed = FALSE, loops = FALSE)

# 3. Two dense communities with sparse between-community links
g_modular <- sample_sbm(
  n = n,
  pref.matrix = matrix(c(0.28, 0.025,
                         0.025, 0.28), nrow = 2, byrow = TRUE),
  block.sizes = c(15, 15),
  directed = FALSE,
  loops = FALSE
)

network_summary <- function(g, name) {
  tibble(
    network = name,
    nodes = vcount(g),
    edges = ecount(g),
    mean_degree = mean(degree(g)),
    clustering = transitivity(g, type = "average"),
    mean_distance = mean_distance(g)
  )
}

summary_df <- bind_rows(
  network_summary(g_ring, "Ring"),
  network_summary(g_random, "Random"),
  network_summary(g_modular, "Modular")
)

print(summary_df)

plot_network <- function(g, title, groups = NULL) {
  coords <- layout_with_fr(g)
  plot(
    g,
    layout = coords,
    vertex.size = 8,
    vertex.label = NA,
    vertex.color = if (is.null(groups)) "steelblue" else groups,
    edge.color = "grey75",
    edge.width = 0.7,
    main = title
  )
}

par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))
plot_network(g_ring, "Ring lattice")
plot_network(g_random, "Random graph")
plot_network(
  g_modular,
  "Modular graph",
  groups = ifelse(seq_len(n) <= 15, "tomato", "goldenrod")
)
par(mfrow = c(1, 1))
```

The ring has local regularity and long paths between distant nodes. The random graph sacrifices much of that local order in exchange for shortcuts. The modular graph creates dense neighbourhoods joined by relatively few bridges. Such differences can shape whether influence remains local, spreads broadly, or becomes trapped within communities.

## R Simulation: Diffusive Dynamics

We can now place a simple nonlinear process on each node. Consider a bistable node model,

$$
\frac{dx_i}{dt}
=
x_i-x_i^3
+
\kappa\sum_{j=1}^{N}A_{ij}(x_j-x_i).
$$

Without coupling, $x_i=-1$ and $x_i=1$ are stable equilibria, while $x_i=0$ is unstable. One can read the two stable states abstractly: low and high activity, inactive and active, or two alternative regimes. Coupling makes neighbouring states matter, so a local perturbation can either decay, remain confined, or recruit a larger part of the network.

``` {r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(igraph)
library(patchwork)

set.seed(42)

simulate_bistable_network <- function(A, x0, kappa = 0.25,
                                      dt = 0.01, tmax = 25) {
  n <- nrow(A)
  steps <- floor(tmax / dt)
  times <- seq(0, tmax, length.out = steps + 1)

  x <- matrix(NA_real_, nrow = steps + 1, ncol = n)
  x[1, ] <- x0

  for (s in seq_len(steps)) {
    current <- x[s, ]
    intrinsic <- current - current^3
    coupling <- kappa * (A %*% current - rowSums(A) * current)
    x[s + 1, ] <- current + dt * (intrinsic + coupling)
  }

  as.data.frame(x) |>
    mutate(time = times, .before = 1) |>
    pivot_longer(
      cols = -time,
      names_to = "node",
      values_to = "state"
    ) |>
    mutate(node = as.integer(gsub("V", "", node)))
}

g <- g_modular
A <- as.matrix(as_adjacency_matrix(g, sparse = FALSE))

# Most nodes begin near the lower stable state.
# Three nodes in the first community are perturbed toward the upper state.
x0 <- rep(-0.9, n)
x0[c(2, 5, 9)] <- 0.9

bistable_df <- simulate_bistable_network(
  A = A,
  x0 = x0,
  kappa = 0.32,
  dt = 0.01,
  tmax = 25
)

node_info <- tibble(
  node = seq_len(n),
  community = ifelse(node <= 15, "Community 1", "Community 2")
)

bistable_df <- left_join(bistable_df, node_info, by = "node")

p_states <- ggplot(
  bistable_df,
  aes(time, state, group = node, color = community)
) +
  geom_line(alpha = 0.55, linewidth = 0.45) +
  scale_color_manual(values = c("Community 1" = "tomato",
                                "Community 2" = "steelblue")) +
  labs(
    title = "Bistable dynamics on a modular network",
    subtitle = "A local perturbation may remain local or spread through coupling",
    x = "Time",
    y = "Node state",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

community_mean <- bistable_df |>
  group_by(time, community) |>
  summarise(mean_state = mean(state), .groups = "drop")

p_means <- ggplot(
  community_mean,
  aes(time, mean_state, color = community)
) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Community 1" = "tomato",
                                "Community 2" = "steelblue")) +
  labs(
    title = "Community-average states",
    subtitle = "Modules can respond differently despite belonging to one network",
    x = "Time",
    y = "Mean state",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_states / p_means
```

The result depends on $\kappa$, initial conditions, and the density of connections between communities. At weak coupling, the perturbation may fade or remain limited to its initial neighbourhood. At stronger coupling, the same local shift can propagate. This is an elementary instance of emergence: the collective outcome cannot be inferred from the isolated bistable equation alone, because it is created by the interaction between node dynamics and graph structure.

## Spreading and Spectral Thresholds

Network coupling does not need to be symmetric or synchronizing. In epidemic models, an edge can represent a possible transmission pathway; in social systems, it can represent exposure, communication, or perceived peer behaviour.

For a simple network SIS model, let $p_i(t)$ denote the probability that node $i$ is infected. A mean-field approximation is

$$
\frac{dp_i}{dt}
=
-\mu p_i
+
\beta(1-p_i)\sum_{j=1}^{N}A_{ij}p_j,
$$

where $\beta$ is a transmission rate and $\mu$ is a recovery rate. Near the disease-free state, linearization suggests that persistence is related to the largest eigenvalue $\lambda_1(A)$ of the adjacency matrix. In the standard network mean-field approximation, the threshold is approximately

$$
\frac{\beta}{\mu}
=
\frac{1}{\lambda_1(A)}.
$$

The interpretation is subtle but important: two populations with the same average number of contacts can have different epidemic thresholds if their connection patterns differ. Large eigenvalues may arise from hubs, dense cores, or particular arrangements of edges. The formula is an approximation rather than a universal law, especially for finite, heterogeneous, or non-Markovian processes, but it illustrates why structure affects collective risk. [iris.joshua-becker](https://iris.joshua-becker.com/lab/spectral-radius-network-spread/)

## R Simulation: Epidemic Spreading

The following simulation implements a stochastic discrete-time SIS process on a modular contact network. An infected node recovers with probability determined by $\mu$; a susceptible node becomes infected with a probability that increases with the number of infected neighbours. The simulation is intentionally simple, but it makes the distinction between individual transitions and network-level prevalence visible.

``` {r}
library(igraph)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(42)

simulate_sis <- function(A, infected0, beta = 0.18, mu = 0.12,
                         steps = 150) {
  n <- nrow(A)
  infected <- matrix(0L, nrow = steps + 1, ncol = n)
  infected[1, ] <- infected0

  for (t in seq_len(steps)) {
    current <- infected[t, ]
    next_state <- current

    for (i in seq_len(n)) {
      infected_neighbors <- sum(A[i, ] * current)

      if (current[i] == 1) {
        next_state[i] <- rbinom(1, size = 1, prob = 1 - mu)
      } else {
        p_infect <- 1 - (1 - beta)^infected_neighbors
        next_state[i] <- rbinom(1, size = 1, prob = p_infect)
      }
    }

    infected[t + 1, ] <- next_state
  }

  as.data.frame(infected) |>
    mutate(time = 0:steps, .before = 1) |>
    pivot_longer(
      cols = -time,
      names_to = "node",
      values_to = "infected"
    ) |>
    mutate(node = as.integer(gsub("V", "", node)))
}

g_epi <- sample_sbm(
  n = 60,
  pref.matrix = matrix(c(0.14, 0.015,
                         0.015, 0.14), nrow = 2, byrow = TRUE),
  block.sizes = c(30, 30),
  directed = FALSE,
  loops = FALSE
)

A_epi <- as.matrix(as_adjacency_matrix(g_epi, sparse = FALSE))

infected0 <- rep(0L, 60)
infected0[c(3, 8, 12)] <- 1L

sis_df <- simulate_sis(
  A = A_epi,
  infected0 = infected0,
  beta = 0.16,
  mu = 0.10,
  steps = 150
)

sis_df <- sis_df |>
  mutate(
    community = ifelse(node <= 30, "Community 1", "Community 2")
  )

prevalence_df <- sis_df |>
  group_by(time, community) |>
  summarise(prevalence = mean(infected), .groups = "drop")

p_prevalence <- ggplot(
  prevalence_df,
  aes(time, prevalence, color = community)
) +
  geom_line(linewidth = 0.85) +
  scale_color_manual(values = c("Community 1" = "tomato",
                                "Community 2" = "steelblue")) +
  labs(
    title = "SIS prevalence in a modular contact network",
    subtitle = "Transmission begins in Community 1 and may cross sparse bridges",
    x = "Discrete time",
    y = "Proportion infected",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

heatmap_df <- sis_df |>
  mutate(
    node = factor(node, levels = rev(seq_len(60))),
    status = factor(infected, levels = c(0, 1),
                    labels = c("Susceptible", "Infected"))
  )

p_heatmap <- ggplot(heatmap_df, aes(time, node, fill = status)) +
  geom_tile() +
  scale_fill_manual(values = c("Susceptible" = "grey95",
                               "Infected" = "firebrick")) +
  labs(
    title = "Individual infection histories",
    subtitle = "Rows are nodes; columns are time points",
    x = "Discrete time",
    y = "Node",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom"
  )

p_prevalence / p_heatmap
```

The simulation will vary from run to run because transmission and recovery are stochastic. That is a feature rather than a nuisance. Even with fixed rates and a fixed graph, early outbreaks may extinguish, persist locally, or cross the sparse inter-community boundary at different times. A deterministic average trajectory can obscure that distribution of possibilities.

## Neural and Clinical Networks

The same mathematics is often imported into neuroscience and psychopathology, where nodes may represent brain regions, neural populations, symptoms, or individuals. Functional brain networks are commonly constructed from statistical dependencies among fMRI, EEG, or MEG signals, while effective-connectivity approaches attempt to characterize directed causal influence. These are different objects and should not be conflated with anatomical connectivity. [europepmc](https://europepmc.org/article/pmc/6582769)

A symptom network can similarly represent estimated relations among insomnia, fatigue, rumination, low mood, avoidance, and concentration problems. Such a graph may be clinically suggestive, but an edge estimated from cross-sectional or time-series data is not automatically a causal pathway. Shared unmeasured causes, measurement overlap, regularization choices, temporal sampling, and nonstationarity can all generate or distort apparent network structure. Clinical network interpretations are most credible when modelling assumptions are explicit and competing models are tested.

Still, the network perspective adds something the single-variable model cannot. A relapse-like transition may not arise because one latent variable crosses a threshold. It may instead reflect mutual reinforcement among several processes: sleep disruption increases fatigue, fatigue reduces activity, reduced activity worsens mood, and worsened mood further disrupts sleep. Whether that loop remains contained may depend on coupling strengths, external inputs, and the presence of bridge variables that connect otherwise separate symptom clusters.

## R Simulation: Kuramoto Synchronization

The final example returns to synchronization. It compares a weakly and strongly coupled set of oscillators on the same network, then plots both individual phases and the order parameter $R(t)$. It is useful to read this as a formal illustration of coherence, not as a literal model of any particular neural rhythm.

``` {r}
library(igraph)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(42)

simulate_kuramoto <- function(A, omega, theta0, K,
                              dt = 0.02, tmax = 30) {
  n <- nrow(A)
  steps <- floor(tmax / dt)
  times <- seq(0, tmax, length.out = steps + 1)

  theta <- matrix(NA_real_, nrow = steps + 1, ncol = n)
  theta[1, ] <- theta0

  for (s in seq_len(steps)) {
    current <- theta[s, ]
    phase_diff <- outer(current, current, FUN = function(a, b) b - a)
    coupling <- rowSums(A * sin(phase_diff))

    theta[s + 1, ] <- current + dt * (omega + K * coupling)
    theta[s + 1, ] <- theta[s + 1, ] %% (2 * pi)
  }

  order_parameter <- apply(theta, 1, function(phases) {
    Mod(mean(exp(1i * phases)))
  })

  list(
    long = as.data.frame(theta) |>
      mutate(time = times, .before = 1) |>
      pivot_longer(
        cols = -time,
        names_to = "node",
        values_to = "phase"
      ) |>
      mutate(node = as.integer(gsub("V", "", node))),
    order = tibble(time = times, R = order_parameter)
  )
}

n <- 40
g_sync <- sample_smallworld(
  dim = 1,
  size = n,
  nei = 2,
  p = 0.12,
  loops = FALSE,
  multiple = FALSE
)

A_sync <- as.matrix(as_adjacency_matrix(g_sync, sparse = FALSE))
omega <- rnorm(n, mean = 1.0, sd = 0.18)
theta0 <- runif(n, min = 0, max = 2 * pi)

weak <- simulate_kuramoto(
  A = A_sync,
  omega = omega,
  theta0 = theta0,
  K = 0.03,
  dt = 0.02,
  tmax = 30
)

strong <- simulate_kuramoto(
  A = A_sync,
  omega = omega,
  theta0 = theta0,
  K = 0.20,
  dt = 0.02,
  tmax = 30
)

phase_df <- bind_rows(
  weak$long |> mutate(condition = "Weak coupling"),
  strong$long |> mutate(condition = "Strong coupling")
)

order_df <- bind_rows(
  weak$order |> mutate(condition = "Weak coupling"),
  strong$order |> mutate(condition = "Strong coupling")
)

p_phase <- ggplot(
  phase_df,
  aes(time, phase, group = node, color = factor(node))
) +
  geom_line(alpha = 0.55, linewidth = 0.35) +
  facet_wrap(~ condition, ncol = 1) +
  scale_color_viridis_d(guide = "none") +
  scale_y_continuous(
    breaks = c(0, pi, 2 * pi),
    labels = c("0", "pi", "2pi")
  ) +
  labs(
    title = "Phase trajectories of coupled oscillators",
    subtitle = "Strong coupling can stabilize phase relationships",
    x = "Time",
    y = "Phase"
  ) +
  theme_minimal(base_size = 11)

p_order <- ggplot(order_df, aes(time, R, color = condition)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Weak coupling" = "grey40",
                                "Strong coupling" = "darkorange3")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Kuramoto order parameter",
    subtitle = "R = 0 indicates dispersion; R = 1 indicates phase coherence",
    x = "Time",
    y = "R(t)",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_phase / p_order
```

Strong coupling need not produce complete synchronization, particularly when intrinsic frequencies differ or the graph is modular. It may instead yield partial synchronization, cluster synchronization, intermittent coherence, or metastable switching between configurations. These intermediate regimes are often more plausible for living systems than either total independence or permanent global lockstep.

## Emergence Is Not a Metaphor

Emergence is sometimes used vaguely to mean “complicated.” In network dynamics, it has a narrower and more useful meaning: a macroscopic pattern arises from local rules and interactions, yet is not a property of the isolated components. A single oscillator does not synchronize; a network of interacting oscillators does. A single susceptible person does not generate an epidemic curve; contact structure and transmission dynamics do.

This does not make high-level phenomena mysterious or exempt from explanation. Quite the opposite. Network models try to identify the mechanisms by which collective states arise. But the mapping from microstructure to macrobehaviour is rarely one-to-one. Similar global patterns can be produced by different combinations of topology, coupling, heterogeneity, delay, noise, and external forcing. Conversely, apparently similar networks can behave quite differently if their coupling rules change.

That limitation is especially relevant in applied settings. A network estimated from correlations is not automatically a dynamical model, and a dynamical model is not automatically an intervention map. Graph-theoretical methods have become widely used in brain-network research, but methodological reviews also emphasize the importance of construction choices and the risk of overinterpreting network measures as direct neurobiological mechanisms. [pmc.ncbi.nlm.nih](https://pmc.ncbi.nlm.nih.gov/articles/PMC6326733/)

## Where This Leaves the Model

The progression from ordinary differential equations to stochastic differential equations and now to networked systems changes the unit of analysis. Instead of one state moving through one landscape, we have many states co-evolving on a structure that can amplify, dampen, delay, or redirect local changes.

The central equation,

$$
\frac{dx_i}{dt}
=
f(x_i)
+
\kappa\sum_{j=1}^{N}A_{ij}g(x_j-x_i),
$$

is only a template, but it forces a useful discipline. Any claim about collective change must state what the nodes are, what edges mean, how influence operates, whether the network changes in time, and which observable would count as synchronization, a cascade, recovery, or instability. Without those commitments, “systems thinking” risks becoming a label rather than a model.

The next natural extension is to combine the present framework with the prior post’s stochastic forcing: networked stochastic differential equations in which local noise, shared noise, coupling, and nonlinear feedback jointly determine whether a perturbation stays local or becomes a system-wide transition.
