
# From Network Structure to Attractors: How Density, Autoregression, and Transitivity Shape Basins of Attraction in Complex Dynamical Systems

**Abstract.** This article develops a unified theoretical account of how three structural properties of a network — global density, edge-level autoregressive persistence, and local transitivity — jointly determine the transition probabilities of a dynamical system, and how those probabilities shape the geometry of attractors and basins of attraction. We move from discrete Markov chains on network state spaces, through continuous-time ordinary differential equations, to potential landscapes, providing mathematical formalism, a worked example model, and complete R code for each section.



## 1. The Fundamental Chain of Causation

The central argument of this article can be stated in one sentence before being unpacked: *network structural features determine transition probabilities; those probabilities define a Markov chain or flow on a state space; and that flow, through its long-run geometry, determines attractors, basin volumes, and transition rates between metastable regions.*  Each term in this chain has a formal meaning, and the connections between them are not metaphorical but mathematically precise. Understanding them requires beginning with what a state space is for a networked system and how transitions across it are governed.[^1][^2]

A dynamical network with $N$ nodes has a state $\mathbf{x}_t$ that lives in some space $\mathcal{X}$. For binary networks, $\mathbf{x}_t \in \{0,1\}^N$ and $|\mathcal{X}| = 2^N$. For continuous networks, $\mathbf{x}_t \in \mathbb{R}^N$. In both cases, an *attractor* $\mathcal{A} \subset \mathcal{X}$ is an invariant set toward which nearby trajectories converge, and a *basin of attraction* $\mathcal{B}(\mathcal{A})$ is the set of all initial conditions whose trajectories eventually reach $\mathcal{A}$. [^1][^3] What determines whether a given region of state space acts as a basin is precisely the transition structure — the rules by which the network moves from one state to the next — and those rules are encoded in the network's parameters. [^1]



## 2. Discrete Network Dynamics: The State Space as a Transition Graph

### 2.1 Wuensche's Framework

Andrew Wuensche's foundational work on discrete dynamical networks makes explicit what is often left implicit in applied modeling: the state space of any finite deterministic network is a directed graph of transitions, and this graph decomposes into a collection of basins of attraction.  Every node (state) in the transition graph either lies on a cycle (an attractor) or lies on a transient tree rooted on a cycle. The attractor's basin is the set of all states that eventually flow into it. Perturbations move a system's state within a basin or, if large enough, across the boundary into another basin.[^2][^3]

This picture is deterministic, but it generalizes naturally to stochastic systems by replacing the single directed edge from each state with a probability distribution over successor states.  The resulting object is a Markov chain on $\mathcal{X}$, and its stationary distribution concentrates on recurrent classes that correspond, in the stochastic sense, to attractors. The basin geometry becomes basin probabilities: the probability that a randomly initiated chain ends in a given recurrent class.[^3][^1]

### 2.2 R Code: Enumerating Basins of a Small Binary Network

The following R code illustrates Wuensche's framework for a small network of $N=6$ binary nodes with a threshold activation rule. The code constructs the full transition graph and identifies attractors by following the deterministic map forward from every initial condition.

```{r}
library(igraph)

set.seed(42)
N <- 6

# ------------------------------------------------------------
# 1) Random signed network
# ------------------------------------------------------------
W <- matrix(sample(c(-1, 1), N * N, replace = TRUE), N, N)
diag(W) <- 0

# ------------------------------------------------------------
# 2) State space: all 2^N configurations in {-1, +1}
# ------------------------------------------------------------
all_states <- as.matrix(expand.grid(rep(list(c(-1L, 1L)), N)))
n_states <- nrow(all_states)

# Fast encoding for state lookup
encode_state <- function(state) {
idx <- ifelse(state == 1L, 1, 0)
sum(idx * 2^((seq_along(idx) - 1)))
}

state_keys <- apply(all_states, 1, encode_state)
state_lookup <- setNames(seq_len(n_states), state_keys)

# ------------------------------------------------------------
# 3) Threshold update rule
# ------------------------------------------------------------
threshold_update <- function(state, W) {
out <- as.integer((W %*% state) > 0)
2L * out - 1L
}

# ------------------------------------------------------------
# 4) Build transition map
# ------------------------------------------------------------
trans <- integer(n_states)
for (i in seq_len(n_states)) {
s_next <- threshold_update(all_states[i, ], W)
trans[i] <- state_lookup[as.character(encode_state(s_next))]
}

# ------------------------------------------------------------
# 5) Find canonical attractor for a start state
# ------------------------------------------------------------
find_attractor <- function(start, trans) {
visited <- integer(0)
current <- start

while (!(current %in% visited)) {
visited <- c(visited, current)
current <- trans[current]
}

cycle_start <- which(visited == current)[1]
cycle <- visited[cycle_start:length(visited)]
min(cycle)
}

attractor_of <- sapply(seq_len(n_states), find_attractor, trans = trans)

# ------------------------------------------------------------
# 6) Basin sizes and attractor states
# ------------------------------------------------------------
basin_sizes <- table(attractor_of)
cat("Attractors (canonical state indices) and basin sizes:\n")
print(basin_sizes)

attr_states <- as.integer(names(basin_sizes))
cat("\nAttractor configurations (each row = one attractor state):\n")
print(all_states[attr_states, , drop = FALSE])
```

Running this on a random weight matrix will typically reveal a small number of fixed-point or short-cycle attractors, each pulling a substantial fraction of the $2^6 = 64$ possible states into its basin.  The key observation is that the weight matrix $W$ — which encodes the network structure — entirely determines this basin partition. Changing the density of $W$ (the fraction of nonzero entries) or the magnitude of edge weights directly reshapes how many attractors exist and how large each basin is.[^1][^2][^3]



## 3. From Network Features to Transition Probabilities: The Formal Model

### 3.1 A Stochastic Edge Process

For dynamic networks where edges themselves evolve, the natural unit of analysis shifts from node states to edge states. Let $X_{t,ij} \in \{0,1\}$ denote the presence or absence of the edge between nodes $i$ and $j$ at time $t$. The network state at time $t$ is the full adjacency matrix $\mathbf{X}_t$. Following Yao and colleagues' autoregressive network model, the conditional transition probability for each edge takes the form

$$
X_{t,ij} \mid \mathcal{F}_{t-1} \sim \mathrm{Bernoulli}\!\left(\gamma^{t-1}_{ij}\right), \quad \gamma^{t-1}_{ij} = \alpha^{t-1}_{ij} + X_{t-1,ij}\!\left(1 - \alpha^{t-1}_{ij} - \beta^{t-1}_{ij}\right),
$$

where $\alpha^{t-1}_{ij}$ is the formation probability (probability that an absent edge appears) and $\beta^{t-1}_{ij}$ is the dissolution probability (probability that a present edge disappears).  When $X_{t-1,ij}=0$, the edge forms with probability $\alpha^{t-1}_{ij}$; when $X_{t-1,ij}=1$, it is retained with probability $1-\beta^{t-1}_{ij}$. The model becomes a full dynamic network autoregression when these formation and dissolution probabilities are themselves functions of lagged network statistics.[^4]

### 3.2 The Three Structural Features

We now encode the three network features of central interest — density, autoregressive persistence, and transitivity — into the formation probability through a logistic linear predictor:

$$
P(X_{t,ij}=1 \mid \mathcal{F}_{t-1}) = \sigma\!\left(a_0 + a_1 D_{t-1} + b_1 X_{t-1,ij} + c\, U_{t-1,ij}\right),
$$

where $\sigma(z) = 1/(1+e^{-z})$ is the logistic function, $D_{t-1} = \frac{1}{N(N-1)}\sum_{k\neq l} X_{t-1,kl}$ is the global network density, $X_{t-1,ij}$ is the autoregressive lag of the focal edge, and $U_{t-1,ij} = \sum_k X_{t-1,ik} X_{t-1,kj}$ is the common-neighbor count, which operationalizes transitivity.  The parameters $a_1$, $b_1$, and $c$ govern the contribution of each structural feature, and their signs and magnitudes jointly determine the long-run dynamical behavior of the network.[^4][^1]

### 3.3 R Code: Simulating the Autoregressive Network Model

```{r}
# ============================================================
# Section 3: Dynamic network simulation
# with empirical basin-fraction estimation
# ============================================================

library(ggplot2)
library(igraph)

set.seed(123)

N <- 10
T <- 300

# ------------------------------------------------------------
# Scenarios
# ------------------------------------------------------------
scenarios <- list(
  list(a0 = -2.0, a1 = 0.0, b1 = 3.0, c = 0.0, label = "Persistence only"),
  list(a0 = -1.5, a1 = 4.0, b1 = 1.5, c = 0.0, label = "Density feedback"),
  list(a0 = -2.0, a1 = 0.0, b1 = 1.5, c = 0.3, label = "Transitivity")
)

sigmoid <- function(z) 1 / (1 + exp(-z))

# ------------------------------------------------------------
# Initialize random undirected network
# ------------------------------------------------------------
init_network <- function(N, p_edge = 0.3) {
  X <- matrix(0, N, N)
  X[upper.tri(X)] <- rbinom(N * (N - 1) / 2, 1, p_edge)
  X <- X + t(X)
  X
}

# ------------------------------------------------------------
# One-step network update
# ------------------------------------------------------------
update_network <- function(X, params) {
  a0 <- params$a0
  a1 <- params$a1
  b1 <- params$b1
  c  <- params$c
  
  N <- nrow(X)
  
  D <- sum(X) / (N * (N - 1))
  U <- X %*% X
  
  eta <- a0 + a1 * D + b1 * X + c * U
  p <- sigmoid(eta)
  p[lower.tri(p, diag = TRUE)] <- 0
  
  X_new <- X
  for (i in 1:(N - 1)) {
    for (j in (i + 1):N) {
      X_new[i, j] <- rbinom(1, 1, p[i, j])
      X_new[j, i] <- X_new[i, j]
    }
  }
  
  X_new
}

# ------------------------------------------------------------
# Simulate one trajectory and record summaries
# ------------------------------------------------------------
simulate_dynnet <- function(params, N, T, init_p = 0.3) {
  X <- init_network(N, p_edge = init_p)
  
  density_ts <- numeric(T)
  clustering_ts <- numeric(T)
  
  for (t in seq_len(T)) {
    X <- update_network(X, params)
    
    density_ts[t] <- sum(X) / (N * (N - 1))
    clustering_ts[t] <- transitivity(
      graph_from_adjacency_matrix(X, mode = "undirected"),
      type = "global"
    )
    
    if (is.nan(clustering_ts[t])) clustering_ts[t] <- 0
  }
  
  data.frame(
    time = 1:T,
    density = density_ts,
    clustering = clustering_ts,
    scenario = params$label
  )
}

# ------------------------------------------------------------
# Classify long-run regime
# ------------------------------------------------------------
classify_regime <- function(density_ts, clustering_ts,
                            density_low = 0.25,
                            density_high = 0.60,
                            clustering_high = 0.35,
                            tail_window = 50) {
  tail_idx <- seq.int(max(1, length(density_ts) - tail_window + 1), length(density_ts))
  
  d_bar <- mean(density_ts[tail_idx], na.rm = TRUE)
  c_bar <- mean(clustering_ts[tail_idx], na.rm = TRUE)
  
  if (d_bar >= density_high) {
    return("high-density")
  } else if (d_bar <= density_low) {
    return("low-density")
  } else if (c_bar >= clustering_high) {
    return("clustered")
  } else {
    return("intermediate")
  }
}

# ------------------------------------------------------------
# Estimate empirical basin fractions
# ------------------------------------------------------------
estimate_basin_fractions <- function(params, N, T,
                                     n_init = 200,
                                     init_ps = seq(0.05, 0.95, by = 0.05),
                                     density_low = 0.25,
                                     density_high = 0.60,
                                     clustering_high = 0.35,
                                     tail_window = 50) {
  regimes <- character(n_init)
  init_density_used <- numeric(n_init)
  
  for (r in seq_len(n_init)) {
    init_p <- sample(init_ps, 1)
    init_density_used[r] <- init_p
    
    sim <- simulate_dynnet(params, N, T, init_p = init_p)
    
    regimes[r] <- classify_regime(
      density_ts = sim$density,
      clustering_ts = sim$clustering,
      density_low = density_low,
      density_high = density_high,
      clustering_high = clustering_high,
      tail_window = tail_window
    )
  }
  
  basin_tab <- prop.table(table(regimes))
  
  basin_df <- data.frame(
    scenario = params$label,
    regime = names(basin_tab),
    fraction = as.numeric(basin_tab)
  )
  
  attr(basin_df, "raw_regimes") <- data.frame(
    replicate = seq_len(n_init),
    init_p = init_density_used,
    regime = regimes,
    scenario = params$label
  )
  
  basin_df
}

# ------------------------------------------------------------
# Run one representative trajectory per scenario
# ------------------------------------------------------------
results_ts <- do.call(
  rbind,
  lapply(scenarios, simulate_dynnet, N = N, T = T, init_p = 0.3)
)

# ------------------------------------------------------------
# Estimate basin fractions for all scenarios
# ------------------------------------------------------------
basin_list <- lapply(
  scenarios,
  estimate_basin_fractions,
  N = N,
  T = T,
  n_init = 300,
  init_ps = seq(0.05, 0.95, by = 0.05),
  density_low = 0.25,
  density_high = 0.60,
  clustering_high = 0.35,
  tail_window = 50
)

basin_results <- do.call(rbind, basin_list)

# Optional: collect raw replicate-level classifications
raw_basin_runs <- do.call(
  rbind,
  lapply(basin_list, function(x) attr(x, "raw_regimes"))
)

# ------------------------------------------------------------
# Print basin estimates
# ------------------------------------------------------------
cat("Estimated empirical basin fractions:\n")
print(basin_results)

# ------------------------------------------------------------
# Plot density trajectories
# ------------------------------------------------------------
p1 <- ggplot(results_ts, aes(x = time, y = density, colour = scenario)) +
  geom_line(alpha = 0.9) +
  labs(
    title = "Network Density Under Different Structural Regimes",
    x = "Time",
    y = "Density",
    colour = "Scenario"
  ) +
  theme_bw(base_size = 13)

print(p1)

# ------------------------------------------------------------
# Plot empirical basin fractions
# ------------------------------------------------------------
p2 <- ggplot(basin_results, aes(x = regime, y = fraction, fill = regime)) +
  geom_col(width = 0.7) +
  facet_wrap(~ scenario) +
  labs(
    title = "Estimated Empirical Basin Fractions",
    x = "Regime",
    y = "Fraction of initial conditions"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")

print(p2)
```

Examining the time series produced by this code reveals the basin intuition directly.  Under persistence alone, the network settles quickly into either a sparse or dense region and rarely leaves. Under positive density feedback ($a_1 > 0$), the network exhibits bistability: it fluctuates between a sparse and a dense attractor, occasionally crossing from one basin to the other. Under transitivity, density is moderate but clustering becomes self-reinforcing, stabilizing tightly connected subgraphs as metastable community structures.[^1][^4]



## 4. Density and the Geometry of Basins

### 4.1 Positive and Negative Density Dependence

Density dependence modifies formation and dissolution probabilities as a function of the current network occupancy. When $a_1 > 0$, high density raises the probability of further edge formation, creating a positive feedback loop that can drive the system toward a dense attractor. When $a_1 < 0$, the feedback is negative, and the system is repelled from extremes toward an intermediate density regime.  The critical mathematical consequence is a bifurcation in the number of fixed points of the implied mean-field dynamics. Let $\bar{D}_t = E[D_t]$ be the expected density at time $t$, which in the mean-field approximation satisfies[^4]

$$
\bar{D}_{t+1} = \sigma(a_0 + a_1 \bar{D}_t + b_1 \bar{D}_t),
$$

a one-dimensional map. When $(a_1 + b_1)$ is large, this map can have three fixed points: two stable (the attractors) and one unstable (the basin boundary).  The system is bistable, and the basin of each attractor corresponds to the set of initial densities on either side of the unstable fixed point. This is the network analogue of a pitchfork or saddle-node bifurcation in nonlinear dynamics.[^5][^6][^7]

### 4.2 R Code: Bifurcation Diagram for Density Dynamics

```{r}
# ============================================================
# Section 4: Mean-field bifurcation diagram for network density
# ============================================================
library(ggplot2)

sigmoid <- function(z) 1 / (1 + exp(-z))

# Fixed point map: D_{t+1} = sigma(a0 + (a1 + b1)*D_t)
iterate_map <- function(D0, a0, slope, n_iter = 200) {
  D <- D0
  for (i in seq_len(n_iter)) D <- sigmoid(a0 + slope * D)
  D
}

# Sweep coupling strength (a1 + b1) and find attractors
slopes  <- seq(-6, 8, length.out = 300)
D0_vals <- seq(0.01, 0.99, length.out = 30)
a0_val  <- -1.5

bifurc <- do.call(rbind, lapply(slopes, function(sl) {
  attractors <- unique(round(sapply(D0_vals, iterate_map,
                                    a0 = a0_val, slope = sl), 3))
  data.frame(slope = sl, D_attractor = attractors)
}))

ggplot(bifurc, aes(x = slope, y = D_attractor)) +
  geom_point(size = 0.5, alpha = 0.5, colour = "steelblue") +
  labs(title = "Bifurcation Diagram: Network Density Attractors vs. Coupling Strength",
       x = expression(a[1] + b[1] ~ "(total coupling)"),
       y = "Attractor density D*") +
  geom_vline(xintercept = 4, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 4.2, y = 0.85, label = "Bistability onset",
           colour = "tomato", hjust = 0, size = 3.5) +
  theme_bw(base_size = 13)
```

The resulting diagram shows clearly how coupling strength controls the number of attractors.  For low coupling, a single intermediate density is the global attractor. As coupling exceeds a critical threshold (the vertical dashed line), the diagram forks into two branches, signifying the emergence of bistability and the appearance of two distinct basins.  This bifurcation is the mathematical signature of a network becoming capable of multistable behavior.[^6][^5]



## 5. Autoregressive Persistence and Basin Depth

### 5.1 Memory as Dynamical Inertia

The autoregressive coefficient $b_1$ controls the probability that an edge retains its current state across one time step. In the edge-level model, this is precisely the definition of a first-order autoregression: the current state is the best predictor of the next state, weighted by $b_1$.  But the dynamical consequences are broader than predictive accuracy. A large $b_1$ deepens basins by reducing the probability of spontaneous state changes, making the system resistant to perturbations and increasing the expected return time to the attractor after small excursions.[^3][^4]

To formalize this, consider the Lyapunov function interpretation. For a discrete-time linear system $\mathbf{x}_{t+1} = A\mathbf{x}_t + \boldsymbol{\varepsilon}_t$, asymptotic stability requires that the spectral radius $\rho(A) = \max_i |\lambda_i(A)| < 1$. [^8] The depth of the basin — operationalized as the rate at which perturbations decay — is governed by $-\log \rho(A)$: larger spectral gap (further below 1) means faster convergence and deeper basin. [^8][^9] In the network autoregression, $b_1$ is the first-order autoregressive coefficient for each edge, and increasing $b_1$ toward 1 moves the effective spectral radius toward the unit circle, flattening the basin and slowing return times toward near-unit-root (near-nonstationary) behavior. [^4]

### 5.2 R Code: Persistence, Basin Depth, and Return Times

```{r}

```

The expected pattern is that higher $b_1$ produces shorter return times, confirming deeper basins.  Each box at a given $b_1$ value summarizes the distribution of return times over repeated perturbation experiments, and the systematic compression of that distribution as $b_1$ increases is precisely the dynamical fingerprint of basin deepening through autoregressive memory.[^2][^3]



## 6. Transitivity, Local Feedback, and Metastability

### 6.1 Triadic Closure as Nonlinear Reinforcement

Transitivity introduces a higher-order feedback mechanism that is qualitatively different from density and autoregressive effects. The common-neighbor term $U_{t-1,ij} = (\mathbf{X}_{t-1}^2)_{ij}$ counts the number of nodes that are currently connected to both $i$ and $j$, and a positive coefficient $c$ makes the formation of the edge $i\text{-}j$ more likely when this count is high.  This is triadic closure, and its dynamical consequence is that dense local clusters become self-reinforcing: once a clique starts to form, the high common-neighbor counts within it accelerate further edge formation inside it, while nodes outside the cluster have low common-neighbor counts and remain disconnected.[^4]

The result is a landscape with multiple metastable regions corresponding to different community structures.  A "community attractor" is a network configuration where connectivity is concentrated within groups and sparse between them. The basin of such an attractor is the set of initial networks from which triadic closure drives the system toward that community structure. Because different initial partitions can lead to different community configurations, transitivity is the primary mechanism by which networks develop *multiple competing basins*, as distinct from bistability (which is governed by density feedback).  Formally, this is related to the existence of multiple ground states in spin-glass physics and multiple memory patterns in Hopfield networks.[^10][^11][^5][^6][^1]

### 6.2 R Code: Transitivity and Community Formation as Metastable Attractors

```{r}

```

Different initializations will converge to different levels of modularity, illustrating path dependence: the initial configuration determines which community basin the system enters.  This is the discrete network equivalent of a system with multiple equilibria separated by unstable saddle points, and it is entirely produced by the transitivity coefficient $c$.[^5][^2][^1]



## 7. Spectral Geometry: Linking Network Structure to Attractor Modes

### 7.1 The Adjacency Eigenstructure

An elegant synthesis of the structural features described above comes from spectral theory. For a network with adjacency matrix $\mathbf{W}$, the eigendecomposition $\mathbf{W} = \mathbf{V}\Lambda\mathbf{V}^{-1}$ reveals the natural modes of collective activity.  A 2023 *Physical Review E* paper demonstrates that in Boolean threshold dynamics, the leading eigenvectors of the adjacency matrix predict which collective states become attractors: modes with large positive eigenvalues correspond to strongly self-reinforcing collective activity patterns.  The spectral structure thus provides an analytical shortcut to identifying the dominant attractors without simulating the full state space.[^12]

In continuous-time linear dynamics, the connection is even more direct. Consider the Hopfield neural network energy function

$$
E(\mathbf{s}) = -\frac{1}{2} \mathbf{s}^\top \mathbf{W} \mathbf{s} + \boldsymbol{\theta}^\top \mathbf{s},
$$

where $\mathbf{s} \in \{-1,1\}^N$ is the state vector and $\boldsymbol{\theta}$ are the thresholds.  This energy function is a Lyapunov function for the synchronous update dynamics, and its minima are the attractors. The number and location of these minima depend on the eigenstructure of $\mathbf{W}$: each sufficiently positive eigenvalue introduces a new energy valley, and the associated eigenvector points toward the attractor configuration.  Network density scales all eigenvalues uniformly; transitivity introduces correlations among rows and columns of $\mathbf{W}$ that can split or merge eigenvalue clusters; and autoregressive self-connections (diagonal entries) shift all eigenvalues by a constant, raising or lowering the global energy surface.[^11][^10][^12]

### 7.2 R Code: Spectral Geometry and Attractor Landscape

```{r}

```

The output table from this code reveals the spectral theory prediction: the number of local energy minima (attractors) tends to grow with the number of large positive eigenvalues.  A clustered network, which has a block structure in $\mathbf{W}$, often exhibits a "spectral gap" — a jump between the first few large eigenvalues and the rest — that corresponds to the number of communities and tends to predict multiple attractors.[^10][^12]



## 8. The Continuous-Time Analogy: ODEs, Lyapunov Functions, and Bifurcations

### 8.1 Network-Coupled ODEs

The discrete framework developed above has a natural continuous-time counterpart. Consider $N$ coupled state variables $\mathbf{x}(t) \in \mathbb{R}^N$ evolving according to

$$
\dot{\mathbf{x}} = f(\mathbf{x}; \mathbf{W}) = -\mathbf{x} + g(\mathbf{W}\mathbf{x} + \boldsymbol{\theta}),
$$

where $g$ is an element-wise nonlinearity, $\mathbf{W}$ is the coupling (network) matrix, and $\boldsymbol{\theta}$ is a threshold or input vector. [^11] This is a continuous Hopfield-type system, and it is also the continuous limit of the VAR framework when the autoregressive coefficient matrix is close to the identity. [^10] Equilibria satisfy $f(\mathbf{x}^*) = \mathbf{0}$, and their stability is governed by the Jacobian $J = Df|_{\mathbf{x}^*} = -I + \mathrm{diag}(g'(\mathbf{W}\mathbf{x}^*))\mathbf{W}$. An equilibrium is locally asymptotically stable if all eigenvalues of $J$ have negative real parts. [^13]

The network features modulate the Jacobian directly. Higher density increases the magnitude of entries in $\mathbf{W}$, which pushes eigenvalues of $J$ toward positive territory and can destabilize equilibria through saddle-node or Hopf bifurcations.  Transitivity (clustered structure in $\mathbf{W}$) introduces block-dominant entries that can create multiple stable equilibria within communities. Autoregressive self-connections are equivalent to the diagonal entries of $\mathbf{W}$, which shift all Jacobian eigenvalues and therefore control whether the origin is a stable point, a saddle, or an unstable repeller.[^14][^7][^11][^10]

### 8.2 R Code: Network-Coupled ODE System and Phase Portraits

```{r}
# ============================================================
# Section 8: Continuous-time network-coupled ODE simulation
# Fixed points, Jacobian stability, and trajectory diagnostics
# ============================================================

library(deSolve)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(nleqslv)

set.seed(2026)

# ------------------------------------------------------------
# Model definition: 12-node network in 3 communities
# ------------------------------------------------------------
N <- 12
group_id <- rep(1:3, each = 4)
node_names <- paste0("x", 1:N)

W <- matrix(0, N, N)

# Within-community excitation
W[1:4, 1:4]     <- 0.85
W[5:8, 5:8]     <- 0.90
W[9:12, 9:12]   <- 0.80
diag(W) <- 0

# Between-community inhibition
W[1:4, 5:12]     <- -0.22
W[5:8, c(1:4, 9:12)] <- -0.22
W[9:12, 1:8]     <- -0.22

# Small heterogeneity in node biases
theta <- seq(-0.05, 0.15, length.out = N)

network_rhs <- function(t, state, parms) {
  x <- state
  W <- parms$W
  th <- parms$theta
  dxdt <- -x + tanh(W %*% x + th)
  list(as.vector(dxdt))
}

network_vec <- function(x, W, theta) {
  as.vector(-x + tanh(W %*% x + theta))
}

network_jac <- function(x, W, theta) {
  u <- W %*% x + theta
  d <- as.vector(1 - tanh(u)^2)
  -diag(N) + diag(d) %*% W
}

# ------------------------------------------------------------
# Trajectory simulation from multiple initial conditions
# ------------------------------------------------------------
n_init <- 40
times <- seq(0, 20, by = 0.05)

traj_list <- vector("list", n_init)
term_list <- vector("list", n_init)

for (k in seq_len(n_init)) {
  x0 <- runif(N, -1.5, 1.5)
  out <- ode(
    y = x0,
    times = times,
    func = network_rhs,
    parms = list(W = W, theta = theta),
    method = "lsoda"
  )
  
  out <- as.data.frame(out)
  colnames(out) <- c("time", node_names)
  out$init <- k
  traj_list[[k]] <- out
  
  xf <- as.numeric(out[nrow(out), node_names])
  term_list[[k]] <- data.frame(
    init = k,
    group1_mean = mean(xf[1:4]),
    group2_mean = mean(xf[5:8]),
    group3_mean = mean(xf[9:12]),
    terminal_energy = -0.5 * as.numeric(t(xf) %*% W %*% xf) - sum(theta * xf)
  )
}

traj_df <- bind_rows(traj_list)
term_df <- bind_rows(term_list)

traj_long <- traj_df %>%
  pivot_longer(cols = all_of(node_names),
               names_to = "node",
               values_to = "value") %>%
  mutate(
    node = factor(node, levels = node_names),
    group = factor(group_id[match(node, node_names)])
  )

# ------------------------------------------------------------
# Numerical fixed-point search
# ------------------------------------------------------------
starts <- 120
fp_list <- vector("list", starts)

for (k in seq_len(starts)) {
  x0 <- runif(N, -2, 2)
  sol <- nleqslv(
    x = x0,
    fn = function(x) network_vec(x, W, theta),
    jac = function(x) network_jac(x, W, theta),
    method = "Newton"
  )
  
  if (sol$termcd %in% c(1, 2)) {
    xstar <- sol$x
    resid <- max(abs(network_vec(xstar, W, theta)))
    if (is.finite(resid) && resid < 1e-7) {
      fp_list[[k]] <- xstar
    }
  }
}

fp_list <- Filter(Negate(is.null), fp_list)

# Deduplicate solutions
fixed_points <- list()
for (x in fp_list) {
  keep <- TRUE
  if (length(fixed_points) > 0) {
    for (y in fixed_points) {
      if (sqrt(sum((x - y)^2)) < 1e-3) {
        keep <- FALSE
        break
      }
    }
  }
  if (keep) fixed_points[[length(fixed_points) + 1]] <- x
}

# ------------------------------------------------------------
# Jacobian stability analysis
# ------------------------------------------------------------
fp_df <- bind_rows(lapply(seq_along(fixed_points), function(i) {
  xstar <- fixed_points[[i]]
  J <- network_jac(xstar, W, theta)
  eig <- eigen(J, only.values = TRUE)$values
  
  data.frame(
    fixed_point = paste0("FP ", i),
    stable = all(Re(eig) < 0),
    max_real_eig = max(Re(eig)),
    group1_mean = mean(xstar[1:4]),
    group2_mean = mean(xstar[5:8]),
    group3_mean = mean(xstar[9:12]),
    terminal_energy = -0.5 * as.numeric(t(xstar) %*% W %*% xstar) - sum(theta * xstar)
  )
}))

fp_long <- bind_rows(lapply(seq_along(fixed_points), function(i) {
  xstar <- fixed_points[[i]]
  data.frame(
    fixed_point = paste0("FP ", i),
    node = factor(node_names, levels = node_names),
    value = xstar,
    stable = all(Re(eigen(network_jac(xstar, W, theta), only.values = TRUE)$values) < 0)
  )
}))

# ------------------------------------------------------------
# Plots
# ------------------------------------------------------------

# 1) Trajectory plot: sample nodes only, to avoid clutter
sample_nodes <- c("x1", "x2", "x3", "x5", "x6", "x7", "x9", "x10", "x11")

p_traj <- traj_long %>%
  filter(node %in% sample_nodes) %>%
  ggplot(aes(x = time, y = value, group = interaction(init, node), color = factor(group))) +
  geom_line(alpha = 0.18, linewidth = 0.45) +
  facet_wrap(~ node, ncol = 3, scales = "free_y") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    strip.background = element_rect(fill = "grey95", color = "grey70"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Network-coupled ODE trajectories",
    subtitle = "Sample of nodes from 40 random initial conditions",
    x = "Time",
    y = "State",
    color = "Community"
  )

# 2) Fixed points projected onto community means
p_fp <- fp_df %>%
  mutate(stability = ifelse(stable, "Stable", "Unstable")) %>%
  ggplot(aes(x = group1_mean, y = group2_mean, color = stability, size = abs(max_real_eig))) +
  geom_point(alpha = 0.9) +
  geom_text(aes(label = fixed_point), nudge_y = 0.03, size = 3, show.legend = FALSE) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Numerically found fixed points",
    subtitle = "Projection onto the first two community means; size indicates |largest real eigenvalue|",
    x = "Group 1 mean",
    y = "Group 2 mean",
    color = "Local stability",
    size = "Eigenvalue"
  )

# 3) Coupling matrix heatmap
W_df <- as.data.frame(as.table(W))
colnames(W_df) <- c("from", "to", "weight")

p_W <- ggplot(W_df, aes(x = to, y = from, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  coord_equal() +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Coupling matrix W",
    subtitle = "Three antagonistic communities with positive within-group and negative between-group coupling",
    x = "To node",
    y = "From node",
    fill = "Weight"
  )

# 4) Terminal summary by attractor-like class
term_df$dominant_group <- factor(
  apply(term_df[, c("group1_mean", "group2_mean", "group3_mean")], 1, which.max),
  levels = 1:3,
  labels = c("Attractor A", "Attractor B", "Attractor C")
)

p_term <- term_df %>%
  pivot_longer(cols = c(group1_mean, group2_mean, group3_mean),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = dominant_group, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Terminal-state summary",
    subtitle = "Group means at the end of simulation",
    x = "Terminal class",
    y = "Mean final value",
    fill = "Metric"
  )

# Combine
(p_W | p_fp) / (p_traj | p_term)

# ------------------------------------------------------------
# Compact interpretation
# ------------------------------------------------------------
fp_summary <- fp_df %>%
  summarise(
    n_fixed_points = n(),
    n_stable = sum(stable),
    n_unstable = sum(!stable)
  )

print(fp_summary)
print(fp_df %>% arrange(desc(stable), max_real_eig))
```

The phase portrait produced here will typically show trajectories from different initial conditions flowing toward one of two attractors, corresponding to the two strongly coupled communities dominating the others.  The boundary between the two flows is the stable manifold of the saddle point, which geometrically separates the two basins.[^13][^11]



## 9. The Potential Landscape View: Connecting to Your Scripts

### 9.1 From Stationary Distributions to Potential Functions

Both scripts you provided earlier take the "potential landscape" approach, which is a natural bridge between probabilistic and dynamical perspectives.  If a Markov chain on a finite state space has a stationary distribution $\pi(\mathbf{x})$, one can define a potential (or energy) function[^15][^16]

$$
U(\mathbf{x}) = -\log \pi(\mathbf{x}),
$$

so that states with high probability have low energy (deep basins) and states with low probability have high energy (basin boundaries and hilltops).  The Ising model does this explicitly: the Boltzmann distribution $\pi(\mathbf{s}) \propto e^{-\beta H(\mathbf{s})}$ makes the energy function the Hamiltonian $H$, and the make_2d_Isingland function in your script computes exactly this potential aggregated over the number of active nodes.[^16][^15]

The VAR script takes the continuous analogue: the stationary distribution of a stable VAR($1$) process $\mathbf{x}_{t+1} = \boldsymbol{\mu} + B\mathbf{x}_t + \boldsymbol{\varepsilon}_t$ is multivariate normal with mean $(I-B)^{-1}\boldsymbol{\mu}$ and covariance $\Sigma_\infty$ satisfying the discrete Lyapunov equation $\Sigma_\infty = B\Sigma_\infty B^\top + \Sigma_\varepsilon$.  The potential over the sum score $S_t = \mathbf{1}^\top \mathbf{x}_t$ is then a Gaussian potential $U(s) = -\log \phi(s; \bar{s}, \sigma_S)$, where $\bar{s} = \mathbf{1}^\top(I-B)^{-1}\boldsymbol{\mu}$ and $\sigma_S^2 = \mathbf{1}^\top\Sigma_\infty \mathbf{1}$.[^16]

### 9.2 R Code: Unifying the Two Scripts — Shared Potential Landscape Framework

```{r}
# ============================================================
# Section 9: Unified potential landscape — Ising and VAR side by side
# ============================================================

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(123)

sigmoid <- function(z) 1 / (1 + exp(-z))

# ------------------------------------------------------------
# Part A: VAR potential landscape
# ------------------------------------------------------------
N <- 9

Beta <- matrix(sample(c(0, 1), N^2, replace = TRUE), N, N) *
  matrix(runif(N^2, 0.08, 0.22), N, N)
diag(Beta) <- runif(N, 0.2, 0.4)

Sigma_eps <- diag(N)

# Ensure stationarity before using the discrete Lyapunov solution
spectral_radius <- function(A) max(Mod(eigen(A, only.values = TRUE)$values))

if (spectral_radius(Beta) >= 0.98) {
  Beta <- Beta / (1.02 * spectral_radius(Beta))
}

lyap_solve <- function(B, S) {
  solve(diag(nrow(B)^2) - kronecker(B, B), as.vector(S))
}

Sigma_inf <- matrix(lyap_solve(Beta, Sigma_eps), N, N)

# Summary diagnostics for the baseline VAR
var_diag <- data.frame(
  spectral_radius = spectral_radius(Beta),
  trace_sigma = sum(diag(Sigma_inf)),
  det_sigma = det(Sigma_inf),
  min_eig_sigma = min(eigen(Sigma_inf, only.values = TRUE)$values),
  max_eig_sigma = max(eigen(Sigma_inf, only.values = TRUE)$values)
)

# Potential curves across coupling multipliers
mult_vals <- c(0.7, 1.0, 1.3)

var_potentials <- bind_rows(lapply(mult_vals, function(mult) {
  Bm <- Beta * mult
  rho <- spectral_radius(Bm)
  if (rho >= 0.98) {
    Bm <- Bm / (1.02 * rho)
  }
  
  Sig_m <- matrix(lyap_solve(Bm, Sigma_eps), N, N)
  mu_sum <- sum(rep(3, N))
  sd_sum <- sqrt(sum(Sig_m))
  x_vals <- seq(-5, 35, length.out = 500)
  U_vals <- -dnorm(x_vals, mean = mu_sum, sd = sd_sum, log = TRUE)
  
  data.frame(
    x = x_vals,
    U = U_vals,
    scenario = paste0("mult = ", mult),
    model = "VAR",
    x_min = mu_sum
  )
}))

var_summary <- var_potentials %>%
  group_by(scenario) %>%
  summarise(
    U_min = min(U),
    U_max = max(U),
    x_at_min = x[which.min(U)],
    .groups = "drop"
  )


# ------------------------------------------------------------
# Part B: Ising-type potential via mean-field approximation
# ------------------------------------------------------------
threshold_vals <- c(-1.0, 0.0, 1.0)
coupling_vals <- c(1.0, 2.0)

ising_grid <- expand.grid(
  threshold = threshold_vals,
  coupling = coupling_vals,
  stringsAsFactors = FALSE
)

ising_list <- vector("list", nrow(ising_grid))

for (i in seq_len(nrow(ising_grid))) {
  h <- ising_grid$threshold[i]
  w <- ising_grid$coupling[i]
  
  k_vals <- 0:9
  raw <- sapply(k_vals, function(k) {
    p <- sigmoid(-h + w * k / 9)
    dbinom(k, 9, p) + 1e-12
  })
  
  prob <- raw / sum(raw)
  
  ising_list[[i]] <- data.frame(
    x = k_vals,
    U = -log(prob),
    scenario = paste0("h = ", h, ", w = ", w),
    threshold = h,
    coupling = w,
    model = "Ising"
  )
}

ising_potentials <- bind_rows(ising_list)

ising_summary <- ising_potentials %>%
  group_by(scenario, threshold, coupling) %>%
  summarise(
    U_min = min(U),
    U_max = max(U),
    k_at_min = x[which.min(U)],
    entropy_like = -sum(exp(-U) * log(exp(-U))),
    .groups = "drop"
  )
# ------------------------------------------------------------
# Additional summary stats
# ------------------------------------------------------------
var_potential_stats <- var_potentials %>%
  group_by(scenario) %>%
  summarise(
    mean_U = mean(U),
    sd_U = sd(U),
    area_U = sum(U),
    .groups = "drop"
  )

ising_potential_stats <- ising_potentials %>%
  group_by(scenario) %>%
  summarise(
    mean_U = mean(U),
    sd_U = sd(U),
    area_U = sum(U),
    .groups = "drop"
  )

# ------------------------------------------------------------
# Plots
# ------------------------------------------------------------
p_var <- ggplot(var_potentials, aes(x = x, y = U, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(data = var_summary, aes(xintercept = x_at_min, colour = scenario),
             linetype = 2, alpha = 0.6, show.legend = FALSE) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank()) +
  labs(
    title = "VAR potential landscape",
    subtitle = "Lower curves indicate smaller negative log-density under the Gaussian sum-score approximation",
    x = "Sum score S",
    y = "U(S) = -log π(S)",
    colour = "Multiplier"
  )

p_ising <- ggplot(ising_potentials, aes(x = x, y = U, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.7) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank()) +
  labs(
    title = "Ising mean-field potential",
    subtitle = "Threshold and coupling shift the discrete energy profile over active-node counts",
    x = "Number of active nodes k",
    y = "U(k)",
    colour = "Setting"
  )

p_var_stats <- var_diag %>%
  mutate(metric = "spectral_radius") %>%
  ggplot(aes(x = metric, y = spectral_radius)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red4") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "VAR stability check",
    subtitle = "Spectral radius below 1 implies a stationary VAR(1) approximation",
    x = "",
    y = "Spectral radius"
  )

p_ising_stats <- ising_summary %>%
  ggplot(aes(x = scenario, y = k_at_min, fill = factor(coupling))) +
  geom_col(width = 0.7) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(
    title = "Ising modal active count",
    subtitle = "Location of the minimum depends on threshold and coupling",
    x = "Scenario",
    y = "k at minimum",
    fill = "Coupling"
  )

# Heatmap of VAR coefficients
Beta_df <- as.data.frame(as.table(Beta))
colnames(Beta_df) <- c("from", "to", "weight")

p_heat <- ggplot(Beta_df, aes(x = to, y = from, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  coord_equal() +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(
    title = "VAR coefficient matrix",
    subtitle = "Sparse random connectivity with diagonal self-coupling",
    x = "To node",
    y = "From node",
    fill = "Beta"
  )

# Combined figure
combined_plot <-
  (p_var | p_ising) /
  (p_var_stats | p_ising_stats) /
  p_heat +
  plot_layout(heights = c(1, 0.8, 1))

print(combined_plot)

# ------------------------------------------------------------
# Tables for reporting
# ------------------------------------------------------------
print(var_diag)
print(var_summary)
print(var_potential_stats)
print(ising_summary)
print(ising_potential_stats)

# ------------------------------------------------------------
# Optional exports
# ------------------------------------------------------------
# ggsave("unified_potential_landscape.png", combined_plot, width = 14, height = 16, dpi = 300)
# write.csv(var_summary, "var_potential_summary.csv", row.names = FALSE)
# write.csv(ising_summary, "ising_potential_summary.csv", row.names = FALSE)
```

This code places the Ising and VAR potential landscapes on a common conceptual footing.  In both cases, the shape of the potential — its number of minima, their depth, and their location — is entirely determined by the network parameters: weights and thresholds in the Ising case, and the coefficient matrix and innovation variance in the VAR case.  The basin of each minimum is the region of state space that flows downhill toward it, and transitions between basins require perturbations large enough to cross the energy barrier (the local maximum between minima).[^6][^15][^5][^16]



## 10. Summary: The Full Causal Chain

The theoretical framework developed across these sections can be organized as a rigorous chain of four propositions.[^10][^1][^4]

**Proposition 1 (Structure → Transitions).** Network density $D$, edge-level persistence $b_1$, and transitivity $c$ enter the logistic transition model as additive predictors. Increasing density feedback ($a_1 > 0$) raises formation probabilities uniformly; increasing persistence ($b_1$) lowers dissolution probability; increasing transitivity ($c$) creates local reinforcement. These changes alter the full $N(N-1)/2$-dimensional transition kernel of the network Markov chain.[^4]

**Proposition 2 (Transitions → Markov Chain Geometry).** The transition kernel defines a Markov chain on the space of network configurations. Its stationary distribution $\pi$ concentrates on recurrent classes (attractors), and the basin structure is determined by the transition probabilities away from each region of state space. Density feedback can create bistability (two recurrent classes); persistence deepens individual basins; transitivity can create multiple community attractors.[^2][^1]

**Proposition 3 (Spectral Structure → Attractor Count).** The eigenvalues of the adjacency or coupling matrix $\mathbf{W}$ constrain which collective modes are self-reinforcing. Large positive eigenvalues correspond to cooperative modes that become attractors; a spectral gap between leading and trailing eigenvalues signals the number of dominant attractor basins. Network structural changes that alter the eigenspectrum (adding density, introducing block structure through transitivity) directly change the number and nature of attractors.[^12][^10]

**Proposition 4 (Continuous Analogy).** In continuous-time network-coupled ODEs, the Jacobian of the vector field at an equilibrium plays the role of the linearized transition matrix. Eigenvalues of the Jacobian with negative real parts confirm local basin structure; bifurcations occur when network parameters push eigenvalues across the imaginary axis, creating or destroying attractors. The AR coefficient in the discrete case and the coupling weights in the ODE case are both mechanisms that control this crossing.[^8][^7][^13]

The practical consequence for research and modeling is that one can read basin structure and transition dynamics off the network parameters without running full simulations: the spectral radius controls global stability, the sign of density dependence controls mono- versus bistability, the magnitude of persistence controls basin depth, and the block structure of the coupling matrix controls the number of competing attractors.  Your Ising and VAR scripts are complementary instantiations of this framework — one discrete with energy minima over configuration space, the other continuous with Gaussian potential over the sum score — both governed by the same underlying logic of network structure shaping long-run dynamical geometry.[^15][^12][^16][^1][^10][^4]
<span style="display:none">[^17][^18][^19][^20][^21][^22][^23][^24][^25][^26][^27][^28][^29][^30][^31][^32][^33][^34]</span>


[^1]: https://www.santafe.edu/research/results/working-papers/basins-of-attraction-in-network-dynamics-a-concept

[^2]: https://www.ddlab.com/downloads/papers/evodevo3.pdf

[^3]: https://users.sussex.ac.uk/~andywu/downloads/papers/2010_basins_of_attraction.pdf

[^4]: http://stats.lse.ac.uk/q.yao/qyao.links/paper/nwDE24.pdf

[^5]: https://www.sciencedirect.com/science/article/pii/S0022249625000690

[^6]: https://link.aps.org/doi/10.1103/PhysRevE.80.031110

[^7]: https://en.wikipedia.org/wiki/Bifurcation_theory

[^8]: http://www.professeurs.polymtl.ca/jerome.le-ny/teaching/CPS/Lec3/4_Stability_notes.pdf

[^9]: https://en.wikipedia.org/wiki/Lyapunov_stability

[^10]: http://www.scholarpedia.org/article/Hopfield_network

[^11]: https://pages.uoregon.edu/meravs/assets/files/Hopfield_Review.pdf

[^12]: https://link.aps.org/doi/10.1103/PhysRevE.108.014402

[^13]: https://link.aps.org/doi/10.1103/PhysRevResearch.5.033139

[^14]: https://pmc.ncbi.nlm.nih.gov/articles/PMC12559148/

[^15]: 250805_IsingPotentialsKendler-2.R

[^16]: 240916_VARpotentials.R

[^17]: https://direct.mit.edu/artl/article/18/3/325/2722/Exploring-Discrete-Dynamics-Andrew-Wuensche-2011

[^18]: https://www.alife.org/wp-content/uploads/2013/08/collections_ECAL93-1150-1159-Wuensche.pdf

[^19]: https://ideas.repec.org/p/wop/safiwp/98-11-101.html

[^20]: https://ideas.repec.org/a/eee/phsmap/v583y2021ics0378437121004817.html

[^21]: https://perso.uclouvain.be/raphael.jungers/sites/default/files/kcfinder/files/pathcomplete.pdf

[^22]: https://link.aps.org/doi/10.1103/PhysRevLett.98.198701

[^23]: https://yonsei.elsevierpure.com/en/publications/approximation-method-to-metastability-an-application-to-nonrevers

[^24]: https://arxiv.org/html/2510.21755v1

[^25]: https://en.wikipedia.org/wiki/Hopfield_network

[^26]: https://www.sciencedirect.com/topics/computer-science/hopfield-neural-networks

[^27]: https://ocw.mit.edu/courses/9-40-introduction-to-neural-computation-spring-2018/d0eb7c14d2fa8ddb6bed2583d2f2962d_MIT9_40S18_Lec20.pdf

[^28]: https://www.igi-global.com/article/optimization-of-hopfield-neural-network-for-improved-pattern-recall-and-storage-using-lyapunov-energy-function-and-hamming-distance/296592

[^29]: https://en.wikipedia.org/wiki/Modern_Hopfield_network

[^30]: https://arxiv.org/html/2605.03557v1

[^31]: https://arxiv.org/html/2505.01218v5

[^32]: https://www.cs.cmu.edu/afs/cs/academic/class/15782-f06/slides/hopfield.pdf

[^33]: https://www.math.uwaterloo.ca/~sacampbe/preprints/siam_shayer.pdf

[^34]: http://webhome.phy.duke.edu/~nb170/teaching/physics567/111-Hopfield.pdf



