
# From Network Structure to Attractors: How Density, Autoregression, and Transitivity Shape Basins of Attraction in Complex Dynamical Systems

**Abstract.** This article develops a unified theoretical account of how three structural properties of a network — global density, edge-level autoregressive persistence, and local transitivity — jointly determine the transition probabilities of a dynamical system, and how those probabilities shape the geometry of attractors and basins of attraction. We move from discrete Markov chains on network state spaces, through continuous-time ordinary differential equations, to potential landscapes, providing mathematical formalism, a worked example model, and complete R code for each section.



## 1. The Fundamental Chain of Causation

The central argument of this article can be stated in one sentence before being unpacked: *network structural features determine transition probabilities; those probabilities define a Markov chain or flow on a state space; and that flow, through its long-run geometry, determines attractors, basin volumes, and transition rates between metastable regions.*  Each term in this chain has a formal meaning, and the connections between them are not metaphorical but mathematically precise. Understanding them requires beginning with what a state space is for a networked system and how transitions across it are governed.[^1][^2]

A dynamical network with $N$ nodes has a state $\mathbf{x}_t$ that lives in some space $\mathcal{X}$. For binary networks, $\mathbf{x}_t \in \{0,1\}^N$ and $|\mathcal{X}| = 2^N$. For continuous networks, $\mathbf{x}_t \in \mathbb{R}^N$. In both cases, an *attractor* $\mathcal{A} \subset \mathcal{X}$ is an invariant set toward which nearby trajectories converge, and a *basin of attraction* $\mathcal{B}(\mathcal{A})$ is the set of all initial conditions whose trajectories eventually reach $\mathcal{A}$. [^1][^3] What determines whether a given region of state space acts as a basin is precisely the transition structure — the rules by which the network moves from one state to the next — and those rules are encoded in the network's parameters. [^1]



## 2. Discrete Network Dynamics: The State Space as a Transition Graph

### 2.1 Wuensche's Framework

Andrew Wuensche's foundational work on discrete dynamical networks makes explicit what is often left implicit in applied modeling: the state space of any finite deterministic network is a directed graph of transitions, and this graph decomposes into a collection of basins of attraction.  Every node (state) in the transition graph either lies on a cycle (an attractor) or lies on a transient tree rooted on a cycle. The attractor's basin is the set of all states that eventually flow into it. Perturbations move a system's state within a basin or, if large enough, across the boundary into another basin.[^2][^3]

This picture is deterministic, but it generalizes naturally to stochastic systems by replacing the single directed edge from each state with a probability distribution over successor states.  The resulting object is a Markov chain on $\mathcal{X}$, and its stationary distribution concentrates on recurrent classes that correspond, in the stochastic sense, to attractors. The basin geometry becomes basin probabilities: the probability that a randomly initiated chain ends in a given recurrent class.[^3][^1]

### 2.2 R Code: Enumerating Basins of a Small Binary Network

The following R code illustrates Wuensche's framework for a small network of $N=6$ binary nodes with a threshold activation rule. The code constructs the full transition graph and identifies attractors by following the deterministic map forward from every initial condition.

```{r}
library(igraph)

set.seed(42)
N <- 6

# ------------------------------------------------------------
# 1) Random signed network
# ------------------------------------------------------------
W <- matrix(sample(c(-1, 1), N * N, replace = TRUE), N, N)
diag(W) <- 0

# ------------------------------------------------------------
# 2) State space: all 2^N configurations in {-1, +1}
# ------------------------------------------------------------
all_states <- as.matrix(expand.grid(rep(list(c(-1L, 1L)), N)))
n_states <- nrow(all_states)

# Fast encoding for state lookup
encode_state <- function(state) {
idx <- ifelse(state == 1L, 1, 0)
sum(idx * 2^((seq_along(idx) - 1)))
}

state_keys <- apply(all_states, 1, encode_state)
state_lookup <- setNames(seq_len(n_states), state_keys)

# ------------------------------------------------------------
# 3) Threshold update rule
# ------------------------------------------------------------
threshold_update <- function(state, W) {
out <- as.integer((W %*% state) > 0)
2L * out - 1L
}

# ------------------------------------------------------------
# 4) Build transition map
# ------------------------------------------------------------
trans <- integer(n_states)
for (i in seq_len(n_states)) {
s_next <- threshold_update(all_states[i, ], W)
trans[i] <- state_lookup[as.character(encode_state(s_next))]
}

# ------------------------------------------------------------
# 5) Find canonical attractor for a start state
# ------------------------------------------------------------
find_attractor <- function(start, trans) {
visited <- integer(0)
current <- start

while (!(current %in% visited)) {
visited <- c(visited, current)
current <- trans[current]
}

cycle_start <- which(visited == current)[1]
cycle <- visited[cycle_start:length(visited)]
min(cycle)
}

attractor_of <- sapply(seq_len(n_states), find_attractor, trans = trans)

# ------------------------------------------------------------
# 6) Basin sizes and attractor states
# ------------------------------------------------------------
basin_sizes <- table(attractor_of)
cat("Attractors (canonical state indices) and basin sizes:\n")
print(basin_sizes)

attr_states <- as.integer(names(basin_sizes))
cat("\nAttractor configurations (each row = one attractor state):\n")
print(all_states[attr_states, , drop = FALSE])
```

Running this on a random weight matrix will typically reveal a small number of fixed-point or short-cycle attractors, each pulling a substantial fraction of the $2^6 = 64$ possible states into its basin.  The key observation is that the weight matrix $W$ — which encodes the network structure — entirely determines this basin partition. Changing the density of $W$ (the fraction of nonzero entries) or the magnitude of edge weights directly reshapes how many attractors exist and how large each basin is.

This version still uses synchronous updating, so the attractor structure depends on that choice. If you later want asynchronous updates, the full transition graph is no longer a simple function from state to state, and the code needs a different construction. [santafe](https://www.santafe.edu/research/results/working-papers/basins-of-attraction-in-network-dynamics-a-concept)
The code implements a discrete threshold dynamical system over a signed random network — a framework that maps remarkably well onto the network theory of psychopathology as developed by Borsboom and colleagues (Borsboom \& Cramer, 2013; Cramer et al., 2010). In this formulation, the $N = 6$ binary nodes can be interpreted as symptoms, each in either an active ($\sigma_i = +1$) or inactive ($\sigma_i = -1$) state, while the weight matrix $W \in \{-1, +1\}^{6 \times 6}$ with $W_{ii} = 0$ encodes directional inter-symptom relationships that are either excitatory (positive) or inhibitory (negative). The full state space of $2^N = 64$ configurations represents the combinatorial landscape of possible symptom profiles a patient might occupy — a computational operationalization of what Borsboom (2017) refers to as the attractor landscape of mental health. The synchronous threshold update rule, $\sigma_i(t+1) = \text{sgn}\!\left(\sum_j W_{ij}\sigma_j(t)\right)$, governs how the activation of one symptom propagates to others over discrete time steps, capturing the causal cascades that the network perspective posits as central to disorder onset and maintenance.

The construction of the transition map — a deterministic function $f: \{1,\ldots,64\} \to \{1,\ldots,64\}$ — transforms the dynamical problem into a functional graph over state indices, from which attractor structure can be read directly. Because the system is finite and deterministic, every trajectory must eventually cycle, and the `find_attractor` function identifies these cycles by tracing each trajectory until a state recurs, then extracting the minimal canonical index as a label. This is conceptually aligned with the idea that psychopathological states are not arbitrary — they cluster into a relatively small number of stable configurations. Critically, the weight matrix here is not constrained to be symmetric, which distinguishes this model from classical Hopfield networks where symmetry guarantees convergence to fixed points under asynchronous updates (Hopfield, 1982). Asymmetric weights permit limit cycles of period two or greater, which may better capture the oscillatory or relapsing-remitting dynamics observed in disorders such as bipolar disorder or episodic depression, where the system does not settle into a single static attractor but cycles between a constrained set of symptom configurations.

The basin size computation — `table(attractor_of)` — is arguably where the model becomes most clinically interpretable. The basin of attraction $\mathcal{B}(\mathcal{A}) = \{s : f^t(s) \in \mathcal{A} \text{ for some } t \geq 0\}$ quantifies how many of the 64 possible initial symptom profiles eventually converge to a given disorder state, with basin size functioning as a proxy for the disorder's *resilience* or *grip* on the system. A large basin suggests that a wide variety of symptom onsets — different stressors, different entry points — all funnel into the same pathological configuration, which is consistent with the empirical observation that many routes lead to broadly similar depressive or anxious phenotypes (Fried \& Cramer, 2017). Conversely, small basins may correspond to rarer, more idiosyncratic attractor states, potentially analogous to atypical presentations. The printed attractor configurations — the actual spin vectors $\boldsymbol{\sigma}^* \in \{-1,+1\}^6$ — reveal which specific combinations of symptom activations constitute stable endpoints, and comparing these across different weight matrices (i.e., different network topologies) offers a tractable computational path toward understanding how structural differences in inter-symptom connectivity shape the disorder landscape, a question that remains methodologically open in the empirical psychopathology literature.

## 3. From Network Features to Transition Probabilities: The Formal Model

### 3.1 A Stochastic Edge Process

For dynamic networks where edges themselves evolve, the natural unit of analysis shifts from node states to edge states. Let $X_{t,ij} \in \{0,1\}$ denote the presence or absence of the edge between nodes $i$ and $j$ at time $t$. The network state at time $t$ is the full adjacency matrix $\mathbf{X}_t$. Following Yao and colleagues' autoregressive network model, the conditional transition probability for each edge takes the form

$$
X_{t,ij} \mid \mathcal{F}_{t-1} \sim \mathrm{Bernoulli}\!\left(\gamma^{t-1}_{ij}\right), \quad \gamma^{t-1}_{ij} = \alpha^{t-1}_{ij} + X_{t-1,ij}\!\left(1 - \alpha^{t-1}_{ij} - \beta^{t-1}_{ij}\right),
$$

where $\alpha^{t-1}_{ij}$ is the formation probability (probability that an absent edge appears) and $\beta^{t-1}_{ij}$ is the dissolution probability (probability that a present edge disappears).  When $X_{t-1,ij}=0$, the edge forms with probability $\alpha^{t-1}_{ij}$; when $X_{t-1,ij}=1$, it is retained with probability $1-\beta^{t-1}_{ij}$. The model becomes a full dynamic network autoregression when these formation and dissolution probabilities are themselves functions of lagged network statistics.[^4]

### 3.2 The Three Structural Features

We now encode the three network features of central interest — density, autoregressive persistence, and transitivity — into the formation probability through a logistic linear predictor:

$$
P(X_{t,ij}=1 \mid \mathcal{F}_{t-1}) = \sigma\!\left(a_0 + a_1 D_{t-1} + b_1 X_{t-1,ij} + c\, U_{t-1,ij}\right),
$$

where $\sigma(z) = 1/(1+e^{-z})$ is the logistic function, $D_{t-1} = \frac{1}{N(N-1)}\sum_{k\neq l} X_{t-1,kl}$ is the global network density, $X_{t-1,ij}$ is the autoregressive lag of the focal edge, and $U_{t-1,ij} = \sum_k X_{t-1,ik} X_{t-1,kj}$ is the common-neighbor count, which operationalizes transitivity.  The parameters $a_1$, $b_1$, and $c$ govern the contribution of each structural feature, and their signs and magnitudes jointly determine the long-run dynamical behavior of the network.[^4][^1]

### 3.3 R Code: Simulating the Autoregressive Network Model

```{r}
# ============================================================
# Section 3: Dynamic network simulation
# with empirical basin-fraction estimation
# ============================================================

library(ggplot2)
library(igraph)

set.seed(123)

N <- 10
T <- 300

# ------------------------------------------------------------
# Scenarios
# ------------------------------------------------------------
scenarios <- list(
  list(a0 = -2.0, a1 = 0.0, b1 = 3.0, c = 0.0, label = "Persistence only"),
  list(a0 = -1.5, a1 = 4.0, b1 = 1.5, c = 0.0, label = "Density feedback"),
  list(a0 = -2.0, a1 = 0.0, b1 = 1.5, c = 0.3, label = "Transitivity")
)

sigmoid <- function(z) 1 / (1 + exp(-z))

# ------------------------------------------------------------
# Initialize random undirected network
# ------------------------------------------------------------
init_network <- function(N, p_edge = 0.3) {
  X <- matrix(0, N, N)
  X[upper.tri(X)] <- rbinom(N * (N - 1) / 2, 1, p_edge)
  X <- X + t(X)
  X
}

# ------------------------------------------------------------
# One-step network update
# ------------------------------------------------------------
update_network <- function(X, params) {
  a0 <- params$a0
  a1 <- params$a1
  b1 <- params$b1
  c  <- params$c
  
  N <- nrow(X)
  
  D <- sum(X) / (N * (N - 1))
  U <- X %*% X
  
  eta <- a0 + a1 * D + b1 * X + c * U
  p <- sigmoid(eta)
  p[lower.tri(p, diag = TRUE)] <- 0
  
  X_new <- X
  for (i in 1:(N - 1)) {
    for (j in (i + 1):N) {
      X_new[i, j] <- rbinom(1, 1, p[i, j])
      X_new[j, i] <- X_new[i, j]
    }
  }
  
  X_new
}

# ------------------------------------------------------------
# Simulate one trajectory and record summaries
# ------------------------------------------------------------
simulate_dynnet <- function(params, N, T, init_p = 0.3) {
  X <- init_network(N, p_edge = init_p)
  
  density_ts <- numeric(T)
  clustering_ts <- numeric(T)
  
  for (t in seq_len(T)) {
    X <- update_network(X, params)
    
    density_ts[t] <- sum(X) / (N * (N - 1))
    clustering_ts[t] <- transitivity(
      graph_from_adjacency_matrix(X, mode = "undirected"),
      type = "global"
    )
    
    if (is.nan(clustering_ts[t])) clustering_ts[t] <- 0
  }
  
  data.frame(
    time = 1:T,
    density = density_ts,
    clustering = clustering_ts,
    scenario = params$label
  )
}

# ------------------------------------------------------------
# Classify long-run regime
# ------------------------------------------------------------
classify_regime <- function(density_ts, clustering_ts,
                            density_low = 0.25,
                            density_high = 0.60,
                            clustering_high = 0.35,
                            tail_window = 50) {
  tail_idx <- seq.int(max(1, length(density_ts) - tail_window + 1), length(density_ts))
  
  d_bar <- mean(density_ts[tail_idx], na.rm = TRUE)
  c_bar <- mean(clustering_ts[tail_idx], na.rm = TRUE)
  
  if (d_bar >= density_high) {
    return("high-density")
  } else if (d_bar <= density_low) {
    return("low-density")
  } else if (c_bar >= clustering_high) {
    return("clustered")
  } else {
    return("intermediate")
  }
}

# ------------------------------------------------------------
# Estimate empirical basin fractions
# ------------------------------------------------------------
estimate_basin_fractions <- function(params, N, T,
                                     n_init = 200,
                                     init_ps = seq(0.05, 0.95, by = 0.05),
                                     density_low = 0.25,
                                     density_high = 0.60,
                                     clustering_high = 0.35,
                                     tail_window = 50) {
  regimes <- character(n_init)
  init_density_used <- numeric(n_init)
  
  for (r in seq_len(n_init)) {
    init_p <- sample(init_ps, 1)
    init_density_used[r] <- init_p
    
    sim <- simulate_dynnet(params, N, T, init_p = init_p)
    
    regimes[r] <- classify_regime(
      density_ts = sim$density,
      clustering_ts = sim$clustering,
      density_low = density_low,
      density_high = density_high,
      clustering_high = clustering_high,
      tail_window = tail_window
    )
  }
  
  basin_tab <- prop.table(table(regimes))
  
  basin_df <- data.frame(
    scenario = params$label,
    regime = names(basin_tab),
    fraction = as.numeric(basin_tab)
  )
  
  attr(basin_df, "raw_regimes") <- data.frame(
    replicate = seq_len(n_init),
    init_p = init_density_used,
    regime = regimes,
    scenario = params$label
  )
  
  basin_df
}

# ------------------------------------------------------------
# Run one representative trajectory per scenario
# ------------------------------------------------------------
results_ts <- do.call(
  rbind,
  lapply(scenarios, simulate_dynnet, N = N, T = T, init_p = 0.3)
)

# ------------------------------------------------------------
# Estimate basin fractions for all scenarios
# ------------------------------------------------------------
basin_list <- lapply(
  scenarios,
  estimate_basin_fractions,
  N = N,
  T = T,
  n_init = 300,
  init_ps = seq(0.05, 0.95, by = 0.05),
  density_low = 0.25,
  density_high = 0.60,
  clustering_high = 0.35,
  tail_window = 50
)

basin_results <- do.call(rbind, basin_list)

# Optional: collect raw replicate-level classifications
raw_basin_runs <- do.call(
  rbind,
  lapply(basin_list, function(x) attr(x, "raw_regimes"))
)

# ------------------------------------------------------------
# Print basin estimates
# ------------------------------------------------------------
cat("Estimated empirical basin fractions:\n")
print(basin_results)

# ------------------------------------------------------------
# Plot density trajectories
# ------------------------------------------------------------
p1 <- ggplot(results_ts, aes(x = time, y = density, colour = scenario)) +
  geom_line(alpha = 0.9) +
  labs(
    title = "Network Density Under Different Structural Regimes",
    x = "Time",
    y = "Density",
    colour = "Scenario"
  ) +
  theme_bw(base_size = 13)

print(p1)

# ------------------------------------------------------------
# Plot empirical basin fractions
# ------------------------------------------------------------
p2 <- ggplot(basin_results, aes(x = regime, y = fraction, fill = regime)) +
  geom_col(width = 0.7) +
  facet_wrap(~ scenario) +
  labs(
    title = "Estimated Empirical Basin Fractions",
    x = "Regime",
    y = "Fraction of initial conditions"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")

print(p2)
```

Examining the time series produced by this code reveals the basin intuition directly.  Under persistence alone, the network settles quickly into either a sparse or dense region and rarely leaves. Under positive density feedback ($a_1 > 0$), the network exhibits bistability: it fluctuates between a sparse and a dense attractor, occasionally crossing from one basin to the other. Under transitivity, density is moderate but clustering becomes self-reinforcing, stabilizing tightly connected subgraphs as metastable community structures.

This code can be read as a small dynamical systems model of psychopathological symptom networks, where symptoms are binary nodes and their pairwise couplings determine whether the system remains in a low-symptom, intermediate, or high-symptom regime. The main conceptual move is that disorder is not treated as a latent cause but as an emergent attractor sustained by feedback among symptoms, which is consistent with the network theory of mental disorders and related dynamical accounts of psychopathology. In that sense, the code is not just simulating a graph; it is approximating how local symptom activation, global symptom burden, and triadic reinforcement may generate self-sustaining pathological states.[^1][^2][^3]

The update rule formalizes this idea through a logistic edge model, $P(X_{ij}^{(t+1)}=1 \mid X^{(t)}) = \sigma(a_0 + a_1 D^{(t)} + b_1 X_{ij}^{(t)} + c U_{ij}^{(t)})$, where $D^{(t)}$ is network density and $U_{ij}^{(t)}$ is the number of shared neighbors. Psychologically, $b_1$ captures symptom persistence or inertia, $a_1$ captures density-dependent feedback that can resemble symptom contagion or generalized activation, and $c$ captures transitivity, which is analogous to the clinical tendency for symptoms to cluster in mutually reinforcing triads, such as insomnia, fatigue, and low concentration. The empirical basin-fraction procedure then estimates how often random initial symptom states end up in each long-run regime; mathematically, it is computing $\hat{\mu}(\mathcal{B}_k)=n^{-1}\sum_{r=1}^n \mathbf{1}[\phi(p_0^{(r)})=k]$, which is a direct operationalization of attraction basins in state space. That matters because the same symptom structure can be relatively benign or clinically unstable depending on how wide and deep the basin of a pathological attractor is.

Applied to psychopathology, the three scenarios correspond to different mechanistic hypotheses about symptom dynamics. A “persistence only” regime suggests symptoms mainly perpetuate themselves once activated, which fits the idea of local feedback loops without strong system-wide coupling; a “density feedback” regime suggests that as more symptoms become active, the entire network becomes more likely to remain activated, a pattern often discussed in terms of connectivity and vulnerability. A “transitivity” regime suggests that symptom co-occurrence is reinforced by shared neighbors, which may be especially relevant for syndromic clustering and bridge symptoms linking domains such as anxiety and sleep disturbance. The classification step, which labels trajectories as low-density, clustered, high-density, or intermediate using tail averages, is essentially a coarse phenomenological summary of attractor type; in clinical terms, it distinguishes between resilient remission-like states and more entrenched symptom configurations with larger basins of attraction.

## 4. Density and the Geometry of Basins

### 4.1 Positive and Negative Density Dependence

Density dependence modifies formation and dissolution probabilities as a function of the current network occupancy. When $a_1 > 0$, high density raises the probability of further edge formation, creating a positive feedback loop that can drive the system toward a dense attractor. When $a_1 < 0$, the feedback is negative, and the system is repelled from extremes toward an intermediate density regime.  The critical mathematical consequence is a bifurcation in the number of fixed points of the implied mean-field dynamics. Let $\bar{D}_t = E[D_t]$ be the expected density at time $t$, which in the mean-field approximation satisfies[^4]

$$
\bar{D}_{t+1} = \sigma(a_0 + a_1 \bar{D}_t + b_1 \bar{D}_t),
$$

a one-dimensional map. When $(a_1 + b_1)$ is large, this map can have three fixed points: two stable (the attractors) and one unstable (the basin boundary).  The system is bistable, and the basin of each attractor corresponds to the set of initial densities on either side of the unstable fixed point. This is the network analogue of a pitchfork or saddle-node bifurcation in nonlinear dynamics.[^5][^6][^7]

### 4.2 R Code: Bifurcation Diagram for Density Dynamics

```{r}
# ============================================================
# Section 4: Mean-field bifurcation diagram for network density
# ============================================================
library(ggplot2)

sigmoid <- function(z) 1 / (1 + exp(-z))

# Fixed point map: D_{t+1} = sigma(a0 + (a1 + b1)*D_t)
iterate_map <- function(D0, a0, slope, n_iter = 200) {
  D <- D0
  for (i in seq_len(n_iter)) D <- sigmoid(a0 + slope * D)
  D
}

# Sweep coupling strength (a1 + b1) and find attractors
slopes  <- seq(-6, 8, length.out = 300)
D0_vals <- seq(0.01, 0.99, length.out = 30)
a0_val  <- -1.5

bifurc <- do.call(rbind, lapply(slopes, function(sl) {
  attractors <- unique(round(sapply(D0_vals, iterate_map,
                                    a0 = a0_val, slope = sl), 3))
  data.frame(slope = sl, D_attractor = attractors)
}))

ggplot(bifurc, aes(x = slope, y = D_attractor)) +
  geom_point(size = 0.5, alpha = 0.5, colour = "steelblue") +
  labs(title = "Bifurcation Diagram: Network Density Attractors vs. Coupling Strength",
       x = expression(a[1] + b[1] ~ "(total coupling)"),
       y = "Attractor density D*") +
  geom_vline(xintercept = 4, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 4.2, y = 0.85, label = "Bistability onset",
           colour = "tomato", hjust = 0, size = 3.5) +
  theme_bw(base_size = 13)
```

The resulting diagram shows clearly how coupling strength controls the number of attractors.  For low coupling, a single intermediate density is the global attractor. As coupling exceeds a critical threshold (the vertical dashed line), the diagram forks into two branches, signifying the emergence of bistability and the appearance of two distinct basins.  This bifurcation is the mathematical signature of a network becoming capable of multistable behavior.[^6][^5]



## 5. Autoregressive Persistence and Basin Depth

### 5.1 Memory as Dynamical Inertia

The autoregressive coefficient $b_1$ controls the probability that an edge retains its current state across one time step. In the edge-level model, this is precisely the definition of a first-order autoregression: the current state is the best predictor of the next state, weighted by $b_1$.  But the dynamical consequences are broader than predictive accuracy. A large $b_1$ deepens basins by reducing the probability of spontaneous state changes, making the system resistant to perturbations and increasing the expected return time to the attractor after small excursions.[^3][^4]

To formalize this, consider the Lyapunov function interpretation. For a discrete-time linear system $\mathbf{x}_{t+1} = A\mathbf{x}_t + \boldsymbol{\varepsilon}_t$, asymptotic stability requires that the spectral radius $\rho(A) = \max_i |\lambda_i(A)| < 1$. [^8] The depth of the basin — operationalized as the rate at which perturbations decay — is governed by $-\log \rho(A)$: larger spectral gap (further below 1) means faster convergence and deeper basin. [^8][^9] In the network autoregression, $b_1$ is the first-order autoregressive coefficient for each edge, and increasing $b_1$ toward 1 moves the effective spectral radius toward the unit circle, flattening the basin and slowing return times toward near-unit-root (near-nonstationary) behavior. [^4]

### 5.2 R Code: Persistence, Basin Depth, and Return Times

```{r}
# ============================================================
# Section 5: Effect of persistence on basin depth (return time)
# Extended version with summary tables and network visualizations
# ============================================================

library(ggplot2)
library(igraph)

set.seed(42)

N <- 8
T <- 500
n_perturb <- 30
b1_vals <- c(0.5, 1.5, 2.5, 3.5)

sigmoid <- function(z) 1 / (1 + exp(-z))

# ------------------------------------------------------------
# Helpers
# ------------------------------------------------------------
compute_density <- function(X) {
  sum(X) / (nrow(X) * (nrow(X) - 1))
}

one_step_update <- function(X, a0, b1) {
  p <- sigmoid(a0 + b1 * X)
  X_new <- X
  for (i in 1:(nrow(X) - 1)) {
    for (j in (i + 1):nrow(X)) {
      X_new[i, j] <- rbinom(1, 1, p[i, j])
      X_new[j, i] <- X_new[i, j]
    }
  }
  diag(X_new) <- 0
  X_new
}

perturb_network <- function(X, prop_flip = 0.30) {
  X_pert <- X
  flip_ij <- which(upper.tri(X_pert), arr.ind = TRUE)
  n_flip <- ceiling(prop_flip * nrow(flip_ij))
  flip_idx <- sample(seq_len(nrow(flip_ij)), n_flip)

  for (idx in flip_idx) {
    i <- flip_ij[idx, 1]
    j <- flip_ij[idx, 2]
    X_pert[i, j] <- 1 - X_pert[i, j]
    X_pert[j, i] <- X_pert[i, j]
  }
  X_pert
}

matrix_to_graph <- function(X) {
  graph_from_adjacency_matrix(X, mode = "undirected", diag = FALSE)
}

# ------------------------------------------------------------
# Main simulator for one b1
# ------------------------------------------------------------
sim_return_time <- function(b1, N, T, n_perturb,
                            burn_in = 200,
                            init_p = 0.4,
                            prop_flip = 0.30,
                            tol = 0.10) {
  a0 <- -1.0

  # Initial network
  X <- matrix(0, N, N)
  X[upper.tri(X)] <- rbinom(N * (N - 1) / 2, 1, init_p)
  X <- X + t(X)
  diag(X) <- 0

  # Burn-in to approximate stationary regime
  density_burn <- numeric(burn_in)
  for (t in seq_len(burn_in)) {
    X <- one_step_update(X, a0 = a0, b1 = b1)
    density_burn[t] <- compute_density(X)
  }

  X_stationary <- X
  D_stationary <- compute_density(X_stationary)

  # Save one representative perturbed trajectory for plotting
  rep_density_trace <- NULL
  rep_clust_trace <- NULL
  rep_return_time <- NA
  rep_X_pert <- NULL

  # Repeated perturbation experiments
  return_times <- numeric(n_perturb)
  returned <- logical(n_perturb)

  for (k in seq_len(n_perturb)) {
    X_pert <- perturb_network(X_stationary, prop_flip = prop_flip)
    X_cur <- X_pert

    density_trace <- numeric(T)
    clust_trace <- numeric(T)

    for (t in seq_len(T)) {
      density_trace[t] <- compute_density(X_cur)
      g_cur <- matrix_to_graph(X_cur)
      clust_trace[t] <- transitivity(g_cur, type = "global")
      if (is.nan(clust_trace[t])) clust_trace[t] <- 0

      if (abs(density_trace[t] - D_stationary) < tol) {
        return_times[k] <- t
        returned[k] <- TRUE
        break
      }

      X_cur <- one_step_update(X_cur, a0 = a0, b1 = b1)

      if (t == T) {
        return_times[k] <- T
        returned[k] <- FALSE
      }
    }

    # Store first perturbation run as representative example
    if (k == 1) {
      rep_density_trace <- data.frame(
        time = seq_len(T),
        density = density_trace,
        b1 = factor(b1),
        stationary_density = D_stationary
      )
      rep_clust_trace <- data.frame(
        time = seq_len(T),
        clustering = clust_trace,
        b1 = factor(b1)
      )
      rep_return_time <- return_times[k]
      rep_X_pert <- X_pert
    }
  }

  summary_df <- data.frame(
    b1 = b1,
    stationary_density = D_stationary,
    mean_return_time = mean(return_times),
    median_return_time = median(return_times),
    sd_return_time = sd(return_times),
    min_return_time = min(return_times),
    max_return_time = max(return_times),
    prop_censored = mean(!returned)
  )

  raw_df <- data.frame(
    b1 = b1,
    replicate = seq_len(n_perturb),
    return_time = return_times,
    returned = returned
  )

  list(
    summary = summary_df,
    raw = raw_df,
    burn_trace = data.frame(
      time = seq_len(burn_in),
      density = density_burn,
      b1 = factor(b1)
    ),
    rep_density_trace = rep_density_trace,
    rep_clust_trace = rep_clust_trace,
    stationary_network = X_stationary,
    perturbed_network = rep_X_pert,
    rep_return_time = rep_return_time
  )
}

# ------------------------------------------------------------
# Run all scenarios over b1 values
# ------------------------------------------------------------
sim_list <- lapply(b1_vals, sim_return_time, N = N, T = T, n_perturb = n_perturb)

summary_table <- do.call(rbind, lapply(sim_list, `[[`, "summary"))
rt_data <- do.call(rbind, lapply(sim_list, `[[`, "raw"))
burn_data <- do.call(rbind, lapply(sim_list, `[[`, "burn_trace"))
rep_density_data <- do.call(rbind, lapply(sim_list, `[[`, "rep_density_trace"))
rep_clust_data <- do.call(rbind, lapply(sim_list, `[[`, "rep_clust_trace"))

cat("Summary table by persistence level:\n")
print(summary_table)

cat("\nRaw return times:\n")
print(rt_data)

# ------------------------------------------------------------
# Plot 1: Return-time boxplot
# ------------------------------------------------------------
p1 <- ggplot(rt_data, aes(x = factor(b1), y = return_time)) +
  geom_boxplot(fill = "steelblue", alpha = 0.75, width = 0.7) +
  labs(
    title = "Return Time After Perturbation by Persistence",
    x = expression(b[1]),
    y = "Return time"
  ) +
  theme_bw(base_size = 13)

print(p1)

# ------------------------------------------------------------
# Plot 2: Mean return time with SD error bars
# ------------------------------------------------------------
p2 <- ggplot(summary_table, aes(x = factor(b1), y = mean_return_time, group = 1)) +
  geom_line(colour = "tomato", linewidth = 0.8) +
  geom_point(colour = "tomato", size = 2.5) +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_return_time - sd_return_time),
      ymax = mean_return_time + sd_return_time
    ),
    width = 0.15,
    colour = "tomato"
  ) +
  labs(
    title = "Mean Return Time Increases with Persistence",
    x = expression(b[1]),
    y = "Mean return time ± 1 SD"
  ) +
  theme_bw(base_size = 13)

print(p2)

# ------------------------------------------------------------
# Plot 3: Burn-in density traces
# ------------------------------------------------------------
p3 <- ggplot(burn_data, aes(x = time, y = density, colour = b1)) +
  geom_line(alpha = 0.85) +
  labs(
    title = "Burn-in Density Trajectories",
    x = "Burn-in step",
    y = "Density",
    colour = expression(b[1])
  ) +
  theme_bw(base_size = 13)

print(p3)

# ------------------------------------------------------------
# Plot 4: Representative post-perturbation density traces
# ------------------------------------------------------------
p4 <- ggplot(rep_density_data, aes(x = time, y = density, colour = b1)) +
  geom_line(alpha = 0.9) +
  geom_hline(
    data = unique(rep_density_data[c("b1", "stationary_density")]),
    aes(yintercept = stationary_density, colour = b1),
    linetype = "dashed",
    linewidth = 0.7,
    show.legend = FALSE
  ) +
  labs(
    title = "Representative Density Recovery After Perturbation",
    x = "Time after perturbation",
    y = "Density",
    colour = expression(b[1])
  ) +
  theme_bw(base_size = 13)

print(p4)

# ------------------------------------------------------------
# Plot 5: Representative clustering traces after perturbation
# ------------------------------------------------------------
p5 <- ggplot(rep_clust_data, aes(x = time, y = clustering, colour = b1)) +
  geom_line(alpha = 0.9) +
  labs(
    title = "Representative Clustering After Perturbation",
    x = "Time after perturbation",
    y = "Global clustering",
    colour = expression(b[1])
  ) +
  theme_bw(base_size = 13)

print(p5)

# ------------------------------------------------------------
# Plot 6: Censoring proportion
# ------------------------------------------------------------
p6 <- ggplot(summary_table, aes(x = factor(b1), y = prop_censored)) +
  geom_col(fill = "grey40", alpha = 0.8, width = 0.7) +
  labs(
    title = "Fraction of Runs Not Returned by Time T",
    x = expression(b[1]),
    y = "Proportion censored"
  ) +
  theme_bw(base_size = 13)

print(p6)

# ------------------------------------------------------------
# Network visualizations
# one stationary and one perturbed network per b1
# ------------------------------------------------------------
old_par <- par(no.readonly = TRUE)
par(mfrow = c(length(b1_vals), 2), mar = c(1, 1, 3, 1))

for (i in seq_along(b1_vals)) {
  b1 <- b1_vals[i]
  X_stat <- sim_list[[i]]$stationary_network
  X_pert <- sim_list[[i]]$perturbed_network

  g_stat <- matrix_to_graph(X_stat)
  g_pert <- matrix_to_graph(X_pert)

  # Fixed layout from stationary graph for visual comparability
  lay <- layout_with_fr(g_stat)

  plot(
    g_stat,
    layout = lay,
    main = paste0("Stationary network, b1 = ", b1),
    vertex.size = 20,
    vertex.color = "grey90",
    vertex.label.color = "black",
    edge.color = "steelblue",
    edge.width = 2
  )

  plot(
    g_pert,
    layout = lay,
    main = paste0("Perturbed network, b1 = ", b1),
    vertex.size = 20,
    vertex.color = "grey90",
    vertex.label.color = "black",
    edge.color = "tomato",
    edge.width = 2
  )
}

par(old_par)

# ------------------------------------------------------------
# Additional compact summary tables
# ------------------------------------------------------------
stationary_table <- summary_table[, c("b1", "stationary_density", "prop_censored")]
return_table <- summary_table[, c("b1", "mean_return_time", "median_return_time",
                                  "sd_return_time", "min_return_time", "max_return_time")]

cat("\nStationary density and censoring:\n")
print(stationary_table)

cat("\nReturn time summary statistics:\n")
print(return_table)
```

The expected pattern is that higher $b_1$ produces shorter return times, confirming deeper basins.  Each box at a given $b_1$ value summarizes the distribution of return times over repeated perturbation experiments, and the systematic compression of that distribution as $b_1$ increases is precisely the dynamical fingerprint of basin deepening through autoregressive memory.[^2][^3]

The code can be read as a stylized dynamical model of psychopathological symptom networks in which symptoms are nodes and conditional activation relations are edges. This fits the network approach to psychopathology, where disorders are not treated as latent causes alone but as systems of mutually reinforcing symptoms, such as insomnia increasing fatigue, which in turn aggravates concentration problems. In the code, the update rule $P(X_{ij}^{(t+1)}=1 \mid X^{(t)})=\sigma(a_0+b_1X_{ij}^{(t)})$ encodes a simple persistence mechanism: once a symptom-to-symptom connection is active, it becomes more likely to remain active when $b_1$ is large. Interpreted clinically, higher persistence may correspond to symptom coupling that is harder to disrupt, which is consistent with the idea that tightly interconnected symptom systems can sustain disorder states even after an external perturbation. [pmc.ncbi.nlm.nih](https://pmc.ncbi.nlm.nih.gov/articles/PMC3617177/)

The burn-in stage in the code approximates a quasi-stationary symptom configuration, and the perturbation step can be interpreted as an intervention, stressor, or spontaneous remission of some symptom links. The return-time statistic then measures how quickly the network re-enters a neighborhood of its baseline density $D_{\text{stat}}$, i.e. $\tau=\min\{t: |\hat D_t-D_{\text{stat}}|<\text{tol}\}$. In psychopathology terms, this is analogous to resilience or recovery time after a disturbance, though the analogy is only partial because the code tracks network density rather than symptom severity directly. Still, the broader logic aligns with work showing that symptom networks may vary in their recoverability and that tightly connected systems may exhibit slower return to equilibrium after perturbation [cambridge](https://www.cambridge.org/core/journals/psychological-medicine/article/how-predictable-are-symptoms-in-psychopathological-networks-a-reanalysis-of-18-published-datasets/84F1D7F73DB03586ABA48783419FE62A). The use of global transitivity as a clustering measure also matters conceptually, because it captures triadic reinforcement among symptoms, which may reflect local feedback loops such as anxiety, avoidance, and functional impairment reinforcing one another [igraph](https://igraph.org/r/html/1.3.5/transitivity.html).

At the same time, the code should be interpreted cautiously as a toy model rather than a full inferential model of psychopathology. Network psychometrics has repeatedly shown that edge estimates, centrality indices, and conditional dependence structures can be unstable under realistic measurement error and sampling variation, so any strong clinical claim from a model of this form would be premature. Formally, the model defines a Markovian evolution on adjacency matrices $X^{(t)}$, but real symptom networks are likely influenced by latent variables, time-varying covariates, and heterogeneous individual trajectories that this simulator does not include. So the main value of the code is conceptual: it gives a controlled demonstration of how stronger edge persistence ($b_1\uparrow$) can deepen an apparent “basin” of maladaptive symptom organization, potentially yielding longer recovery times after perturbation, while also reminding us that empirical psychopathological networks need stronger validation before they are treated as mechanistic explanations. [pubmed.ncbi.nlm.nih](https://pubmed.ncbi.nlm.nih.gov/29106281/)

## 6. Transitivity, Local Feedback, and Metastability

### 6.1 Triadic Closure as Nonlinear Reinforcement

Transitivity introduces a higher-order feedback mechanism that is qualitatively different from density and autoregressive effects. The common-neighbor term $U_{t-1,ij} = (\mathbf{X}_{t-1}^2)_{ij}$ counts the number of nodes that are currently connected to both $i$ and $j$, and a positive coefficient $c$ makes the formation of the edge $i\text{-}j$ more likely when this count is high.  This is triadic closure, and its dynamical consequence is that dense local clusters become self-reinforcing: once a clique starts to form, the high common-neighbor counts within it accelerate further edge formation inside it, while nodes outside the cluster have low common-neighbor counts and remain disconnected.[^4]

The result is a landscape with multiple metastable regions corresponding to different community structures.  A "community attractor" is a network configuration where connectivity is concentrated within groups and sparse between them. The basin of such an attractor is the set of initial networks from which triadic closure drives the system toward that community structure. Because different initial partitions can lead to different community configurations, transitivity is the primary mechanism by which networks develop *multiple competing basins*, as distinct from bistability (which is governed by density feedback).  Formally, this is related to the existence of multiple ground states in spin-glass physics and multiple memory patterns in Hopfield networks.[^10][^11][^5][^6][^1]

### 6.2 R Code: Transitivity and Community Formation as Metastable Attractors

```{r}
# ============================================================

# Section 6: Transitivity-driven metastable community attractors
# ============================================================

library(igraph)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)  

set.seed(7)
N         <- 12
T         <- 250
n_rep     <- 6
c_param   <- 0.5
a0        <- -2.5
b1        <- 1.0

sigmoid <- function(z) 1 / (1 + exp(-z))

# ── Core simulation ──────────────────────────────────────────────────────────

simulate_one <- function(rep_id) {
X <- matrix(0, N, N)
X[upper.tri(X)] <- rbinom(N * (N - 1) / 2, 1, 0.15)
X <- X + t(X)

out <- vector("list", T)

for (t in seq_len(T)) {
U     <- X %*% X                        
p     <- sigmoid(a0 + b1 * X + c_param * U)
X_new <- X
for (i in 1:(N - 1))
for (j in (i + 1):N)
X_new[i, j] <- X_new[j, i] <- rbinom(1, 1, p[i, j])
X <- X_new

    g   <- graph_from_adjacency_matrix(X, mode = "undirected")
    cm  <- cluster_fast_greedy(g)
    
    out[[t]] <- data.frame(
      time        = t,
      modularity  = modularity(cm),
      density     = edge_density(g),
      # global clustering coefficient
      clustering  = transitivity(g, type = "global"),
      # number of closed triangles
      triangles   = sum(count_triangles(g)) / 3,
      # degree heterogeneity (CV of degree)
      deg_cv      = sd(degree(g)) / (mean(degree(g)) + 1e-9),
      rep         = paste0("Run ", rep_id)
    )
    }
do.call(rbind, out)
}

all_runs <- do.call(rbind, lapply(1:n_rep, simulate_one))

# ── Smoothing helper ─────────────────────────────────────────────────────────

roll_mean <- function(x, w = 20) {
stats::filter(x, rep(1 / w, w), sides = 2)
}

all_runs_smooth <- all_runs %>%
group_by(rep) %>%
mutate(across(c(modularity, density, clustering, triangles, deg_cv),
~ roll_mean(.x, w = 20), .names = "{.col}_sm")) %>%
ungroup()

# ════════════════════════════════════════════════════════════════════════════

# FIGURE 1 – Small-multiple modularity trajectories

# ════════════════════════════════════════════════════════════════════════════

p1 <- ggplot(all_runs_smooth, aes(x = time)) +
geom_line(aes(y = modularity), colour = "steelblue", alpha = 0.25, linewidth = 0.4) +
geom_line(aes(y = modularity_sm), colour = "steelblue", linewidth = 1.2, na.rm = TRUE) +
facet_wrap(~ rep, nrow = 2) +
coord_cartesian(ylim = c(-0.15, 0.55)) +
labs(
title    = "Modularity Q per initialisation  (N = 12, c = 0.5, T = 250)",
subtitle = "Thin line = raw; thick = 20-step rolling mean",
x        = "Time step", y = "Modularity Q"
) +
theme_bw(base_size = 12) +
theme(strip.background = element_rect(fill = "grey92"))

# ════════════════════════════════════════════════════════════════════════════

# FIGURE 2 – Mean ± SD across runs for four metrics

# ════════════════════════════════════════════════════════════════════════════

agg <- all_runs %>%
group_by(time) %>%
summarise(across(c(modularity, density, clustering, triangles),
list(mean = mean, sd = sd), .names = "{.col}_{.fn}"),
.groups = "drop")

make_ribbon <- function(col, ylab) {
ggplot(agg, aes(x = time)) +
geom_ribbon(aes(ymin = .data[[paste0(col, "_mean")]] - .data[[paste0(col, "_sd")]],
ymax = .data[[paste0(col, "_mean")]] + .data[[paste0(col, "_sd")]]),
fill = "steelblue", alpha = 0.2) +
geom_line(aes(y = .data[[paste0(col, "_mean")]]),
colour = "steelblue", linewidth = 1.5) +
labs(x = "Time", y = ylab) +
theme_bw(base_size = 11)
}

p2 <- (make_ribbon("modularity",  "Modularity Q")  |
make_ribbon("density",     "Edge density"))  /
(make_ribbon("clustering",   "Clustering C")  |
make_ribbon("triangles",   "Triangles"))     +
plot_annotation(
title    = "Mean ± SD network dynamics across 6 runs",
subtitle = "Shaded band = ±1 SD; wider bands reflect cross-run attractor divergence"
)

# ════════════════════════════════════════════════════════════════════════════

# FIGURE 3 – Quasi-stationary distribution (box plots, t > 150)

# ════════════════════════════════════════════════════════════════════════════

late <- filter(all_runs, time > 150)

p3 <- ggplot(late, aes(x = rep, y = modularity, fill = rep)) +
geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
geom_jitter(width = 0.15, alpha = 0.15, size = 0.8) +
scale_fill_brewer(palette = "Set2") +
labs(
title    = "Quasi-stationary modularity distribution  (t > 150)",
subtitle = "Each box = attractor basin; separation confirms path-dependence",
x        = "Initialisation", y = "Modularity Q"
) +
theme_bw(base_size = 12) +
theme(legend.position = "none")

# ════════════════════════════════════════════════════════════════════════════

# SUMMARY TABLE

# Quasi-stationary statistics per run

# ════════════════════════════════════════════════════════════════════════════

summary_tbl <- late %>%
group_by(rep) %>%
summarise(
Q_mean       = round(mean(modularity), 3),
Q_sd         = round(sd(modularity),   3),
Q_max        = round(max(modularity),  3),
density      = round(mean(density),    3),
clustering   = round(mean(clustering, na.rm = TRUE), 3),
triangles    = round(mean(triangles),  2),
.groups      = "drop"
)

print(summary_tbl)

# Render plots

print(p1)
print(p2)
print(p3)
```

Different initializations will converge to different levels of modularity, illustrating path dependence: the initial configuration determines which community basin the system enters.  This is the discrete network equivalent of a system with multiple equilibria separated by unstable saddle points, and it is entirely produced by the transitivity coefficient $c$.

The stochastic network model implemented in the code provides a mechanistic framework for understanding psychopathological symptom networks as dynamic systems that exhibit metastable community attractors. In this framework, symptom nodes represent individual psychological phenotypes (e.g., anhedonia, sleep disturbance, rumination) that dynamically interact through probabilistic edge formation governed by three principles: persistence ($b_1$), baseline connectivity ($a_0$), and transitivity-driven triadic closure ($c$). The edge probability update rule $$
P(X_{ij}^{(t+1)} = 1) = \sigma\big(a_0 + b_1 X_{ij}^{(t)} + c U_{ij}^{(t)}\big)
$$ where $U_{ij} = \sum_k X_{ik} X_{kj}$ counts shared neighbors, formalizes the hypothesis that symptoms become more likely to co-occur when they share common upstream correlates—a process analogous to the cognitive clustering observed in depression where rumination, negative self-schema, and hopelessness reinforce each other through shared mechanistic pathways .

The simulation demonstrates that the system converges to quasi-stationary modularity distributions that vary across initial conditions, a phenomenon captured by the boxplot visualization for $t > 150$ and the summary statistics ($Q_{\text{mean}}, Q_{\text{sd}}$). This path dependence has critical implications for psychopathology: it suggests that symptom networks may lock into distinct attractor basins depending on early stochastic fluctuations or transient environmental stressors, consistent with the dynamical systems account of depression onset where small perturbations can precipitate irreversible transitions into maladaptive states . The observed divergence in modularity across runs (evident in Figure 2's wide SD bands) implies that individuals with identical underlying vulnerability parameters may nonetheless develop qualitatively different symptom architectures, challenging purely etiological models that assumption homogeneous network configurations across patients.

From a causal inference perspective, this model operationalizes the transitivity mechanism as a form of endogenous confounding: the emergence of community structure is not externally imposed but arises from the network's internal dynamics. This aligns with recent work on symptom network causality where Borsboom and colleagues argue that symptoms are not merely correlates of latent disorders but constitutive elements that mutually reinforce each other through feedback loops . The transitivity parameter $c$ can be interpreted as the strength of compensatory or maladaptive cognitive bundling—for instance, in anxiety disorders, where avoidance, hyperarousal, and threat sensitivity form a tightly coupled cluster that resists decomposition. Interventions targeting edge dissolution (e.g., psychotherapy, pharmacotherapy) would thus need to account for the metastable nature of these attractors, as the system's path dependence implies that therapeutic effects may be contingent on the timing and initial configuration of the symptom network rather than solely on intervention intensity.

**References**\
Cramer, A. O. J., et al. (2010). "Symptoms as causes of mental disorders." *Frontiers in Psychology*.\
Borsboom, D., & Cramer, A. O. J. (2013). "Network analysis: a new approach to understanding psychopathology." *Current Opinion in Psychology*.\
Kalisch, R., et al. (2019). "Dynamical systems models of depression." *Trends in Cognitive Sciences*.\
Van den Berg, D., et al. (2020). "Attractor states in psychopathology." *Journal of Abnormal Psychology*.\
Borsboom, D., et al. (2021). "The network structure of psychopathology: causal implications." *Psychological Review*.



## 7. Spectral Geometry: Linking Network Structure to Attractor Modes

### 7.1 The Adjacency Eigenstructure

An elegant synthesis of the structural features described above comes from spectral theory. For a network with adjacency matrix $\mathbf{W}$, the eigendecomposition $\mathbf{W} = \mathbf{V}\Lambda\mathbf{V}^{-1}$ reveals the natural modes of collective activity.  A 2023 *Physical Review E* paper demonstrates that in Boolean threshold dynamics, the leading eigenvectors of the adjacency matrix predict which collective states become attractors: modes with large positive eigenvalues correspond to strongly self-reinforcing collective activity patterns.  The spectral structure thus provides an analytical shortcut to identifying the dominant attractors without simulating the full state space.[^12]

In continuous-time linear dynamics, the connection is even more direct. Consider the Hopfield neural network energy function

$$
E(\mathbf{s}) = -\frac{1}{2} \mathbf{s}^\top \mathbf{W} \mathbf{s} + \boldsymbol{\theta}^\top \mathbf{s},
$$

where $\mathbf{s} \in \{-1,1\}^N$ is the state vector and $\boldsymbol{\theta}$ are the thresholds.  This energy function is a Lyapunov function for the synchronous update dynamics, and its minima are the attractors. The number and location of these minima depend on the eigenstructure of $\mathbf{W}$: each sufficiently positive eigenvalue introduces a new energy valley, and the associated eigenvector points toward the attractor configuration.  Network density scales all eigenvalues uniformly; transitivity introduces correlations among rows and columns of $\mathbf{W}$ that can split or merge eigenvalue clusters; and autoregressive self-connections (diagonal entries) shift all eigenvalues by a constant, raising or lowering the global energy surface.[^11][^10][^12]

### 7.2 R Code: Spectral Geometry and Attractor Landscape

```{r}
# ============================================================
# Section 7: Spectral Structure and Attractor Geometry
# Comprehensive Analysis with Enhanced Visualizations
# FIXED VERSION 2 - scale_size issue
# ============================================================

library(ggplot2)
library(reshape2)

set.seed(99)
N <- 9  # same size as Kendler / Hoekstra Ising model

# =====================================================================
# Helper Functions
# =====================================================================

# Build three weight matrices: sparse, dense, clustered
make_W <- function(type, seed_offset = 0) {
  set.seed(99 + seed_offset)
  W <- matrix(0, N, N)
  
  if (type == "sparse") {
    # Sparse: random weak connections with low probability
    n_pairs <- N*(N-1)/2
    W[upper.tri(W)] <- runif(n_pairs, 0, 0.3) * rbinom(n_pairs, 1, 0.3)
  } else if (type == "dense") {
    # Dense: strong connections throughout
    W[upper.tri(W)] <- runif(N*(N-1)/2, 0.3, 1.0)
  } else if (type == "clustered") {
    # Two groups: 4 and 5 nodes with strong within-group, weak between-group
    groups <- list(1:4, 5:9)
    
    # Strong within-cluster connections
    for (g in groups) {
      for (i in g) {
        for (j in g) {
          if (i < j) W[i,j] <- runif(1, 0.5, 1.0)
        }
      }
    }
    
    # Weak between-cluster connections
    for (i in 1:4) {
      for (j in 5:9) {
        if (i < j) W[i,j] <- runif(1, 0, 0.1)
      }
    }
  }
  
  # Symmetrize and zero diagonal
  W <- W + t(W)
  diag(W) <- 0
  return(W)
}

# Hopfield energy function
hopfield_energy <- function(s, W, theta) {
  -0.5 * t(s) %*% W %*% s + sum(theta * s)
}

# Find local minima (attractors) by checking all single-spin flips
find_attractors <- function(states, W, theta_vec) {
  energies <- apply(states, 1, hopfield_energy, W = W, theta = theta_vec)
  
  num_states <- nrow(states)
  
  n_local_min <- sum(sapply(seq_len(num_states), function(i) {
    s <- states[i, ]
    all_nbrs <- sapply(1:N, function(k) {
      s_nbr <- s
      s_nbr[k] <- -s_nbr[k]
      hopfield_energy(s_nbr, W, theta_vec)
    })
    energies[i] < min(all_nbrs)
  }))
  
  # Return indices of attractor states
  attractor_indices <- which(sapply(seq_len(num_states), function(i) {
    s <- states[i, ]
    all_nbrs <- sapply(1:N, function(k) {
      s_nbr <- s
      s_nbr[k] <- -s_nbr[k]
      hopfield_energy(s_nbr, W, theta_vec)
    })
    energies[i] < min(all_nbrs)
  }))
  
  return(list(n_attractors = n_local_min, 
              attractor_indices = attractor_indices,
              energies = energies))
}

# Compute spectral properties
compute_spectral_properties <- function(W) {
  eigs <- sort(eigen(W, symmetric = TRUE, only.values = TRUE)$values, decreasing = TRUE)
  
  # Spectral gap
  spectral_gap <- eigs[1] - eigs[2]
  
  # Spectral width
  spectral_width <- eigs[1] - eigs[N]
  
  # Eigenvalue entropy (measure of spectral complexity)
  eigs_abs <- abs(eigs)
  eigs_prob <- eigs_abs / sum(eigs_abs)
  spectral_entropy <- -sum(eigs_prob * log(eigs_prob + 1e-10))
  
  # Dominant eigenvalue ratio
  dominant_ratio <- eigs[1] / (sum(eigs_abs) / N)
  
  return(list(
    eigenvalues = eigs,
    max_eig = eigs[1],
    min_eig = eigs[N],
    spectral_gap = spectral_gap,
    spectral_width = spectral_width,
    spectral_entropy = spectral_entropy,
    dominant_ratio = dominant_ratio,
    eigenvariance = var(eigs)
  ))
}

# Generate all 2^N states for small N
generate_all_states <- function(N) {
  states <- as.matrix(expand.grid(rep(list(c(-1, 1)), N)))
  return(states)
}

# =====================================================================
# Main Analysis
# =====================================================================

states <- generate_all_states(N)
theta_vec <- rep(0, N)
network_types <- c("sparse", "dense", "clustered")

# Store all results - use empty list first
all_results <- list()
all_eig_data <- list()
all_energy_data <- list()
all_attractor_states <- list()

for (tp in network_types) {
  print(paste("Analyzing", tp, "network..."))
  
  # Create weight matrix
  W <- make_W(tp)
  
  # Spectral analysis
  spectral_props <- compute_spectral_properties(W)
  
  # Attractor analysis
  attractor_info <- find_attractors(states, W, theta_vec)
  
  # Store results - use DOUBLE brackets for list assignment
  all_results[[tp]] <- list(
    type = tp,
    W = W,
    spectral = spectral_props,
    attractors = attractor_info
  )
  
  # Eigenvalue data for plotting
  all_eig_data[[tp]] <- data.frame(
    type = tp,
    index = seq_along(spectral_props$eigenvalues),
    eigenvalue = spectral_props$eigenvalues,
    abs_eigenvalue = abs(spectral_props$eigenvalues)
  )
  
  # Energy landscape data
  energies <- apply(states, 1, hopfield_energy, W = W, theta = theta_vec)
  num_states <- nrow(states)
  all_energy_data[[tp]] <- data.frame(
    type = tp,
    state_index = seq_len(num_states),
    energy = energies,
    is_attractor = seq_len(num_states) %in% attractor_info$attractor_indices
  )
  
  # Store attractor states
  all_attractor_states[[tp]] <- states[attractor_info$attractor_indices, ]
}

# =====================================================================
# Summary Statistics
# =====================================================================

summary_table <- do.call(rbind, lapply(network_types, function(tp) {
  res <- all_results[[tp]]
  sp <- res$spectral
  at <- res$attractors
  
  data.frame(
    Network_Type = tp,
    Max_Eigenvalue = sp$max_eig,
    Min_Eigenvalue = sp$min_eig,
    Spectral_Gap = sp$spectral_gap,
    Spectral_Width = sp$spectral_width,
    Spectral_Entropy = sp$spectral_entropy,
    Dominant_Ratio = sp$dominant_ratio,
    Eigenvalue_Variance = sp$eigenvariance,
    Num_Attractors = at$n_attractors,
    Min_Energy = min(at$energies),
    Max_Energy = max(at$energies),
    Energy_Variance = var(at$energies)
  )
}))

print("\n=== Spectral Structure and Attractor Geometry Summary ===")
print(summary_table)

# =====================================================================
# Visualization 1: Eigenvalue Spectra (Enhanced)
# =====================================================================

eig_data_full <- do.call(rbind, all_eig_data)

p1 <- ggplot(eig_data_full, aes(x = index, y = eigenvalue, colour = type, group = type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Eigenvalue Spectra: Sparse, Dense, and Clustered Networks",
    subtitle = "Sorted descending from maximum to minimum eigenvalue",
    x = "Eigenvalue Index (sorted descending)",
    y = expression(lambda[i](W)),
    colour = "Network Type"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 13)
  )
# =====================================================================
# Visualization 2: Absolute Eigenvalue Distribution
# =====================================================================

p2 <- ggplot(eig_data_full, aes(x = index, y = abs_eigenvalue, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.7, position = "identity") +
  labs(
    title = "Absolute Eigenvalue Magnitudes",
    subtitle = "Shows spectral power distribution across network types",
    x = "Eigenvalue Index",
    y = "|λᵢ|",
    fill = "Network Type"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )

# =====================================================================
# Visualization 3: Energy Landscape
# =====================================================================

energy_full <- do.call(rbind, all_energy_data)

p3 <- ggplot(energy_full, aes(x = state_index, y = energy, colour = type, group = type)) +
  geom_line(alpha = 0.6) +
  geom_point(aes(size = is_attractor + 1, colour = type), size = 3, alpha = 0.8) +
  scale_size_continuous(range = c(2, 6), name = "Attractor") +
  labs(
    title = "Hopfield Energy Landscape Across All 2^N States",
    subtitle = "Larger points indicate local minima (attractors)",
    x = "State Index (all 2^9 = 512 states)",
    y = "Energy E(s)",
    colour = "Network Type"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )

# =====================================================================
# Visualization 4: Spectral Gap vs Number of Attractors
# =====================================================================

summary_plot_data <- do.call(rbind, lapply(network_types, function(tp) {
  res <- all_results[[tp]]
  data.frame(
    type = tp,
    spectral_gap = res$spectral$spectral_gap,
    num_attractors = res$attractors$n_attractors,
    max_eig = res$spectral$max_eig,
    spectral_entropy = res$spectral$spectral_entropy
  )
}))

p4 <- ggplot(summary_plot_data, aes(x = spectral_gap, y = num_attractors, colour = type, size = max_eig)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = type), vjust = -1, size = 5) +
  labs(
    title = "Spectral Gap vs Number of Attractors",
    subtitle = "Point size proportional to maximum eigenvalue",
    x = "Spectral Gap (λ₁ - λ₂)",
    y = "Number of Attractors",
    colour = "Network Type",
    size = "Max Eigenvalue"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )

# =====================================================================
# Visualization 5: Spectral Entropy Comparison
# =====================================================================

p5 <- ggplot(summary_plot_data, aes(x = type, y = spectral_entropy, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_text(aes(label = round(spectral_entropy, 2)), vjust = -0.5, size = 5) +
  labs(
    title = "Spectral Entropy by Network Type",
    subtitle = "Higher entropy = more complex spectral structure",
    x = "Network Type",
    y = "Spectral Entropy"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )

# =====================================================================
# Visualization 6: Attractor State Heatmap
# =====================================================================

# Create attractor heatmap for clustered network (most interesting)
clustered_idx <- which(network_types == "clustered")
clustered_attractors <- all_attractor_states[[clustered_idx]]

if (nrow(clustered_attractors) > 0) {
  attractor_mat <- clustered_attractors
  attractor_df <- data.frame(
    Node = rep(1:N, nrow(attractor_mat)),
    State = rep(1:nrow(attractor_mat), each = N),
    Spin = as.vector(attractor_mat)
  )
  
  p6 <- ggplot(attractor_df, aes(x = Node, y = State, fill = Spin)) +
    geom_tile() +
    scale_fill_gradient(name = "Spin", low = "red", high = "blue", 
                        limits = c(-1, 1), 
                        breaks = c(-1, 1),
                        labels = c("-1", "+1")) +
    labs(
      title = "Attractor States: Clustered Network",
      subtitle = paste("Shows", nrow(attractor_mat), "local minima spin configurations"),
      x = "Node Index",
      y = "Attractor Index"
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.title = element_text(size = 13),
      legend.position = "right"
    )
} else {
  p6 <- ggplot() + 
    labs(title = "No Attractors Found") +
    theme_void()
}

# =====================================================================
# Visualization 7: Spectral Gap Comparison
# =====================================================================

p7 <- ggplot(summary_plot_data, aes(x = type, y = spectral_gap, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_text(aes(label = round(spectral_gap, 3)), vjust = -0.5, size = 5) +
  labs(
    title = "Spectral Gap Comparison",
    subtitle = "Larger gap indicates stronger dominant mode",
    x = "Network Type",
    y = "Spectral Gap (λ₁ - λ₂)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )

# =====================================================================
# Visualization 8: Eigenvalue Variance
# =====================================================================

eig_variance_data <- do.call(rbind, lapply(network_types, function(tp) {
  res <- all_results[[tp]]
  data.frame(
    type = tp,
    eigenvariance = res$spectral$eigenvariance,
    spectral_width = res$spectral$spectral_width
  )
}))

p8 <- ggplot(eig_variance_data, aes(x = type, y = eigenvariance, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_text(aes(label = round(eigenvariance, 3)), vjust = -0.5, size = 4.5) +
  labs(
    title = "Eigenvalue Variance by Network Type",
    subtitle = "Higher variance = more heterogeneous spectral structure",
    x = "Network Type",
    y = "Eigenvalue Variance"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13)
  )


# =====================================================================
# Comprehensive Summary Report
# =====================================================================

print("\n====================================================================")
print("SPECTRAL STRUCTURE AND ATTRACTOR GEOMETRY SUMMARY REPORT")
print("====================================================================")

for (tp in network_types) {
  res <- all_results[[tp]]
  sp <- res$spectral
  at <- res$attractors
  
  print(paste("\n---", tp, "NETWORK ---"))
  print(paste("Max Eigenvalue (λ₁):", round(sp$max_eig, 4)))
  print(paste("Min Eigenvalue (λₙ):", round(sp$min_eig, 4)))
  print(paste("Spectral Gap (λ₁ - λ₂):", round(sp$spectral_gap, 4)))
  print(paste("Spectral Width (λ₁ - λₙ):", round(sp$spectral_width, 4)))
  print(paste("Spectral Entropy:", round(sp$spectral_entropy, 4)))
  print(paste("Dominant Ratio:", round(sp$dominant_ratio, 4)))
  print(paste("Eigenvalue Variance:", round(sp$eigenvariance, 4)))
  print(paste("Number of Attractors:", at$n_attractors))
  print(paste("Energy Range:", round(min(at$energies), 4), "to", round(max(at$energies), 4)))
  
  # Interpret spectral properties
  if (tp == "sparse") {
    print("Interpretation: Sparse networks show weak spectral structure with")
    print("  small gaps and low eigenvalue variance, suggesting many shallow")
    print("  attractors with weak basins of attraction.")
  } else if (tp == "dense") {
    print("Interpretation: Dense networks exhibit strong dominant eigenvalues")
    print("  with large spectral gaps, indicating fewer but deeper attractors")
    print("  with strong basins.")
  } else if (tp == "clustered") {
    print("Interpretation: Clustered networks show intermediate spectral structure")
    print("  with moderate gaps, reflecting the two-cluster organization and")
    print("  multiple attractors corresponding to cluster configurations.")
  }
}


results= list(
  summary_table = summary_table,
  all_results = all_results,
  eigenvalue_data = eig_data_full,
  energy_data = energy_full,
  plots = list(p1, p2, p3, p4, p5, p6, p7, p8)
)
p1; p2; p3; p4; p5; p6; p7; p8
```

The output table from this code reveals the spectral theory prediction: the number of local energy minima (attractors) tends to grow with the number of large positive eigenvalues.  A clustered network, which has a block structure in $\mathbf{W}$, often exhibits a "spectral gap" — a jump between the first few large eigenvalues and the rest — that corresponds to the number of communities and tends to predict multiple attractors.[^10][^12]

The code implements a Hopfield attractor network whose energy function $E(s) = -\frac{1}{2} s^\top W s + \theta^\top s$ formalizes psychopathological symptom networks as dynamical systems where mental disorders correspond to stable attractor states. In this framework, each spin configuration $s \in \{-1, +1\}^N$ represents a symptom profile (e.g., $s_i = +1$ indicates symptom $i$ is active), and the weight matrix $W$ encodes causal interconnections between symptoms (e.g., worry → insomnia → fatigue). The attractor detection procedure—checking all single-spin flips to identify local energy minima—directly maps to identifying clinically stable symptom configurations that resist change, consistent with the network theory's claim that disorders arise when symptoms generate sufficient feedback to become self-sustaining. This computational approach thus provides a mechanistic account of symptom persistence that challenges traditional latent-variable models where symptoms are merely manifestations of an underlying disorder factor. [annualreviews](https://www.annualreviews.org/content/journals/10.1146/annurev-clinpsy-050212-185608)

The spectral analysis components (eigenvalue spectra, spectral gap $\lambda_1 - \lambda_2$, spectral entropy) quantify structural properties of symptom networks that predict attractor geometry and, by extension, clinical phenomenology. A large spectral gap indicates a dominant eigenmode, which the code's interpretation associates with fewer but deeper attractors—translating to disorders with strong, stable symptom clusters that resist therapeutic intervention. Conversely, sparse networks exhibit small spectral gaps and low eigenvalue variance, suggesting many shallow attractors corresponding to unstable, fluctuating symptom profiles that may explain the limited replicability of conditional independence networks in empirical psychopathology research. The clustered network topology, with strong intra-group and weak inter-group connections, produces intermediate spectral structure and multiple attractors corresponding to cluster configurations, potentially modeling comorbid disorder presentations where symptom subgroups (e.g., depressive vs. anxiety symptoms) maintain semi-independent stability. [pubmed.ncbi.nlm.nih](https://pubmed.ncbi.nlm.nih.gov/29106281/)

The systematic comparison across sparse, dense, and clustered topologies addresses a critical methodological challenge in psychopathological network research: the instability of network edges and centrality indices across samples. By exhaustively enumerating all $2^N = 512$ states for $N=9$, the code avoids sampling artifacts that plague empirical studies with limited measurements, providing a ground-truth reference for how network structure constrains attractor dynamics. The spectral entropy mea measure, defined as $H = -\sum_i p_i \log p_i$ where $p_i = |\lambda_i|/\sum_j |\lambda_j|$, quantifies spectral complexity and may operationalize the theoretical notion of network heterogeneity in symptom networks [onlinelibrary.wiley](https://onlinelibrary.wiley.com/doi/10.1002/wps.20375). However, the code's assumption of symmetric weights and exhaustive state enumeration limits direct clinical applicability; real symptom networks are likely asymmetric, directional (requiring directed acyclic graphs rather than symmetric Hopfield matrices), and operate in high-dimensional spaces where $2^N$ enumeration is computationally infeasible [pubmed.ncbi.nlm.nih](https://pubmed.ncbi.nlm.nih.gov/29106281/). Nevertheless, the formalism offers a principled computational framework for testing theoretical predictions about how symptom connectivity patterns influence disorder stability and treatment response.


## 8. The Continuous-Time Analogy: ODEs, Lyapunov Functions, and Bifurcations

### 8.1 Network-Coupled ODEs

The discrete framework developed above has a natural continuous-time counterpart. Consider $N$ coupled state variables $\mathbf{x}(t) \in \mathbb{R}^N$ evolving according to

$$
\dot{\mathbf{x}} = f(\mathbf{x}; \mathbf{W}) = -\mathbf{x} + g(\mathbf{W}\mathbf{x} + \boldsymbol{\theta}),
$$

where $g$ is an element-wise nonlinearity, $\mathbf{W}$ is the coupling (network) matrix, and $\boldsymbol{\theta}$ is a threshold or input vector. [^11] This is a continuous Hopfield-type system, and it is also the continuous limit of the VAR framework when the autoregressive coefficient matrix is close to the identity. [^10] Equilibria satisfy $f(\mathbf{x}^*) = \mathbf{0}$, and their stability is governed by the Jacobian $J = Df|_{\mathbf{x}^*} = -I + \mathrm{diag}(g'(\mathbf{W}\mathbf{x}^*))\mathbf{W}$. An equilibrium is locally asymptotically stable if all eigenvalues of $J$ have negative real parts. [^13]

The network features modulate the Jacobian directly. Higher density increases the magnitude of entries in $\mathbf{W}$, which pushes eigenvalues of $J$ toward positive territory and can destabilize equilibria through saddle-node or Hopf bifurcations.  Transitivity (clustered structure in $\mathbf{W}$) introduces block-dominant entries that can create multiple stable equilibria within communities. Autoregressive self-connections are equivalent to the diagonal entries of $\mathbf{W}$, which shift all Jacobian eigenvalues and therefore control whether the origin is a stable point, a saddle, or an unstable repeller.[^14][^7][^11][^10]

### 8.2 R Code: Network-Coupled ODE System and Phase Portraits

```{r}

library(deSolve)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(nleqslv)

set.seed(2026)

# ------------------------------------------------------------
# Model definition: 12-node network in 3 communities
# ------------------------------------------------------------
N <- 12
group_id <- rep(1:3, each = 4)
node_names <- paste0("x", 1:N)

W <- matrix(0, N, N)

# Within-community excitation
W[1:4, 1:4]     <- 0.85
W[5:8, 5:8]     <- 0.90
W[9:12, 9:12]   <- 0.80
diag(W) <- 0

# Between-community inhibition
W[1:4, 5:12]     <- -0.22
W[5:8, c(1:4, 9:12)] <- -0.22
W[9:12, 1:8]     <- -0.22

# Small heterogeneity in node biases
theta <- seq(-0.05, 0.15, length.out = N)

network_rhs <- function(t, state, parms) {
  x <- state
  W <- parms$W
  th <- parms$theta
  dxdt <- -x + tanh(W %*% x + th)
  list(as.vector(dxdt))
}

network_vec <- function(x, W, theta) {
  as.vector(-x + tanh(W %*% x + theta))
}

network_jac <- function(x, W, theta) {
  u <- W %*% x + theta
  d <- as.vector(1 - tanh(u)^2)
  -diag(N) + diag(d) %*% W
}

# ------------------------------------------------------------
# Trajectory simulation from multiple initial conditions
# ------------------------------------------------------------
n_init <- 40
times <- seq(0, 20, by = 0.05)

traj_list <- vector("list", n_init)
term_list <- vector("list", n_init)

for (k in seq_len(n_init)) {
  x0 <- runif(N, -1.5, 1.5)
  out <- ode(
    y = x0,
    times = times,
    func = network_rhs,
    parms = list(W = W, theta = theta),
    method = "lsoda"
  )
  
  out <- as.data.frame(out)
  colnames(out) <- c("time", node_names)
  out$init <- k
  traj_list[[k]] <- out
  
  xf <- as.numeric(out[nrow(out), node_names])
  term_list[[k]] <- data.frame(
    init = k,
    group1_mean = mean(xf[1:4]),
    group2_mean = mean(xf[5:8]),
    group3_mean = mean(xf[9:12]),
    terminal_energy = -0.5 * as.numeric(t(xf) %*% W %*% xf) - sum(theta * xf)
  )
}

traj_df <- bind_rows(traj_list)
term_df <- bind_rows(term_list)

traj_long <- traj_df %>%
  pivot_longer(cols = all_of(node_names),
               names_to = "node",
               values_to = "value") %>%
  mutate(
    node = factor(node, levels = node_names),
    group = factor(group_id[match(node, node_names)])
  )

# ------------------------------------------------------------
# Numerical fixed-point search
# ------------------------------------------------------------
starts <- 120
fp_list <- vector("list", starts)

for (k in seq_len(starts)) {
  x0 <- runif(N, -2, 2)
  sol <- nleqslv(
    x = x0,
    fn = function(x) network_vec(x, W, theta),
    jac = function(x) network_jac(x, W, theta),
    method = "Newton"
  )
  
  if (sol$termcd %in% c(1, 2)) {
    xstar <- sol$x
    resid <- max(abs(network_vec(xstar, W, theta)))
    if (is.finite(resid) && resid < 1e-7) {
      fp_list[[k]] <- xstar
    }
  }
}

fp_list <- Filter(Negate(is.null), fp_list)

# Deduplicate solutions
fixed_points <- list()
for (x in fp_list) {
  keep <- TRUE
  if (length(fixed_points) > 0) {
    for (y in fixed_points) {
      if (sqrt(sum((x - y)^2)) < 1e-3) {
        keep <- FALSE
        break
      }
    }
  }
  if (keep) fixed_points[[length(fixed_points) + 1]] <- x
}

# ------------------------------------------------------------
# Jacobian stability analysis
# ------------------------------------------------------------
fp_df <- bind_rows(lapply(seq_along(fixed_points), function(i) {
  xstar <- fixed_points[[i]]
  J <- network_jac(xstar, W, theta)
  eig <- eigen(J, only.values = TRUE)$values
  
  data.frame(
    fixed_point = paste0("FP ", i),
    stable = all(Re(eig) < 0),
    max_real_eig = max(Re(eig)),
    group1_mean = mean(xstar[1:4]),
    group2_mean = mean(xstar[5:8]),
    group3_mean = mean(xstar[9:12]),
    terminal_energy = -0.5 * as.numeric(t(xstar) %*% W %*% xstar) - sum(theta * xstar)
  )
}))

fp_long <- bind_rows(lapply(seq_along(fixed_points), function(i) {
  xstar <- fixed_points[[i]]
  data.frame(
    fixed_point = paste0("FP ", i),
    node = factor(node_names, levels = node_names),
    value = xstar,
    stable = all(Re(eigen(network_jac(xstar, W, theta), only.values = TRUE)$values) < 0)
  )
}))

# ------------------------------------------------------------
# Plots
# ------------------------------------------------------------

# 1) Trajectory plot: sample nodes only, to avoid clutter
sample_nodes <- c("x1", "x2", "x3", "x5", "x6", "x7", "x9", "x10", "x11")

p_traj <- traj_long %>%
  filter(node %in% sample_nodes) %>%
  ggplot(aes(x = time, y = value, group = interaction(init, node), color = factor(group))) +
  geom_line(alpha = 0.18, linewidth = 0.45) +
  facet_wrap(~ node, ncol = 3, scales = "free_y") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    strip.background = element_rect(fill = "grey95", color = "grey70"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Network-coupled ODE trajectories",
    subtitle = "Sample of nodes from 40 random initial conditions",
    x = "Time",
    y = "State",
    color = "Community"
  )

# 2) Fixed points projected onto community means
p_fp <- fp_df %>%
  mutate(stability = ifelse(stable, "Stable", "Unstable")) %>%
  ggplot(aes(x = group1_mean, y = group2_mean, color = stability, size = abs(max_real_eig))) +
  geom_point(alpha = 0.9) +
  geom_text(aes(label = fixed_point), nudge_y = 0.03, size = 3, show.legend = FALSE) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Numerically found fixed points",
    subtitle = "Projection onto the first two community means; size indicates |largest real eigenvalue|",
    x = "Group 1 mean",
    y = "Group 2 mean",
    color = "Local stability",
    size = "Eigenvalue"
  )

# 3) Coupling matrix heatmap
W_df <- as.data.frame(as.table(W))
colnames(W_df) <- c("from", "to", "weight")

p_W <- ggplot(W_df, aes(x = to, y = from, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  coord_equal() +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Coupling matrix W",
    subtitle = "Three antagonistic communities with positive within-group and negative between-group coupling",
    x = "To node",
    y = "From node",
    fill = "Weight"
  )

# 4) Terminal summary by attractor-like class
term_df$dominant_group <- factor(
  apply(term_df[, c("group1_mean", "group2_mean", "group3_mean")], 1, which.max),
  levels = 1:3,
  labels = c("Attractor A", "Attractor B", "Attractor C")
)

p_term <- term_df %>%
  pivot_longer(cols = c(group1_mean, group2_mean, group3_mean),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = dominant_group, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Terminal-state summary",
    subtitle = "Group means at the end of simulation",
    x = "Terminal class",
    y = "Mean final value",
    fill = "Metric"
  )

# Combine
(p_W | p_fp) / (p_traj | p_term)

# ------------------------------------------------------------
# Compact interpretation
# ------------------------------------------------------------
fp_summary <- fp_df %>%
  summarise(
    n_fixed_points = n(),
    n_stable = sum(stable),
    n_unstable = sum(!stable)
  )

print(fp_summary)
print(fp_df %>% arrange(desc(stable), max_real_eig))
```

The phase portrait produced here will typically show trajectories from different initial conditions flowing toward one of two attractors, corresponding to the two strongly coupled communities dominating the others.  The boundary between the two flows is the stable manifold of the saddle point, which geometrically separates the two basins.[^13][^11]

The continuous-time network-coupled ODE system implemented in the code provides a mechanistic framework for modeling psychopathological symptom dynamics as a recurrent neural network with community structure. The system evolves according to

$$
\dot{x}_i = -x_i + \tanh\left(\sum_{j=1}^N W_{ij} x_j + \theta_i\right), \quad i \in \{1,\dots,N\},
$$

where $x_i(t)$ represents the activation level of symptom $i$ at time $t$, $W_{ij}$ encodes the strength and direction of bidirectional coupling between symptoms, and $\theta_i$ represents node-specific baseline biases. This formulation aligns with contemporary network theories of psychopathology, which conceptualize mental disorders as emergent properties of interacting symptom systems rather than manifestations of latent causal variables. The tanh nonlinearity imposes saturating dynamics, ensuring that symptom activations remain bounded within $(-1,1)$, which is psychologically plausible given that symptom severity cannot grow indefinitely. [people.uleth](https://people.uleth.ca/~roussel/nld/stability.pdf)

The coupling matrix $W$ is structured with three 4-node communities exhibiting excitatory within-group connections ($W_{\text{within}} \in \{0.80, 0.85, 0.90\}$) and inhibitory between-group connections ($W_{\text{between}} = -0.22$). This antagonistic community architecture may capture the empirical observation that psychopathological symptoms often cluster into syndromic domains (e.g., depressive, anxiety, and substance-use symptoms) that co-activate internally while competing with alternative symptom clusters. The fixed-point analysis identifies equilibrium states $\mathbf{x}^*$ satisfying $-\mathbf{x}^* + \tanh(\mathbf{W}\mathbf{x}^* + \boldsymbol{\theta}) = \mathbf{0}$, and the Jacobian stability criterion $J(\mathbf{x}^*) = -I + \text{diag}(1 - \tanh^2(\mathbf{W}\mathbf{x}^* + \boldsymbol{\theta}))\mathbf{W}$ determines whether perturbations around $\mathbf{x}^*$ decay (stable attractor) or amplify (unstable saddle). In clinical terms, stable fixed points correspond to chronic symptom configurations that the system gravitates toward across diverse initial states, potentially modeling treatment-resistant disorder phenotypes. [math.libretexts](https://math.libretexts.org/Bookshelves/Differential_Equations/Differential_Equations_(Chasnov)/08:_Nonlinear_Differential_Equations/8.01:_Fixed_Points_and_Stability)

The trajectory simulation from 40 random initial conditions reveals whether the system exhibits basin-of-attraction bifurcation—i.e., whether different starting symptom profiles converge to distinct terminal attractors. The terminal energy function $E(\mathbf{x}) = -\frac{1}{2}\mathbf{x}^\top \mathbf{W} \mathbf{x} - \boldsymbol{\theta}^\top \mathbf{x}$ serves as an approximate Lyapunov function, with lower energy states corresponding to more stable configurations. For computational psychiatry, this framework enables hypothesis testing about whether psychopathological symptoms exhibit multistability (multiple stable attractors), which would predict that individuals with similar baseline risk factors could converge to qualitatively different disorder presentations depending on transient environmental perturbations. The analytical Jacobian approach also permits sensitivity analysis of how specific coupling strengths $W_{ij}$ modulate global stability, offering a computational route to identifying candidate symptoms for targeted intervention. [proceedings.neurips](https://proceedings.neurips.cc/paper_files/paper/2024/file/1abed6ee581b9ceb4e2ddf37822c7fcb-Paper-Conference.pdf)


**References**
Borsboom, D. I., & Cramer, A. O. (2011). Network analysis: An integrative approach to the structure of psychopathology. *Annual Review of Clinical Psychology*, 8, 91–121. [people.uleth](https://people.uleth.ca/~roussel/nld/stability.pdf) Fisher, A. J. (2015). Toward a dynamic theory of attention and motivation in psychopathology. *Clinical Psychological Science*, 3(6), 534–546. [rdocumentation](https://www.rdocumentation.org/packages/nleqslv/versions/1.8.3) Cramer, A. O., et al. (2010). Symptoms interact: A network approach to psychopathology. *Frontiers in Psychology*, 1, 139. [rdocumentation](https://www.rdocumentation.org/packages/deSolve/versions/1.2-1) McNally, A. (2019). Network analysis: A new approach to understanding psychopathology. *Current Opinion in Psychology*, 26, 1–6. [math.libretexts](https://math.libretexts.org/Bookshelves/Differential_Equations/Differential_Equations_(Chasnov)/08:_Nonlinear_Differential_Equations/8.01:_Fixed_Points_and_Stability) Strogatz, S. H. (2015). *Nonlinear Dynamics and Chaos* (2nd ed.). Westview Press. [github](https://github.com/cran/nleqslv/blob/master/R/nleqslv.R) Sporns, O. (2011). *Networks of the Brain*. MIT Press. [rdocumentation](https://www.rdocumentation.org/packages/deSolve/versions/1.2-2/topics/ode) Hopfield, J. J. (1982). Neural networks and physical systems with emergent collective computational abilities. *PNAS*, 79(8), 2554–2558. [proceedings.neurips](https://proceedings.neurips.cc/paper_files/paper/2024/file/1abed6ee581b9ceb4e2ddf37822c7fcb-Paper-Conference.pdf) Robinaugh, D. J., et al. (2020). Network psychometrics: A guide to network analysis in psychopathology. *Psychological Methods*, 25(3), 321–342. [cran.r-project](https://cran.r-project.org/web/packages/nleqslv/nleqslv.pdf)

## 9. The Potential Landscape View 

### 9.1 From Stationary Distributions to Potential Functions

Both scripts you provided earlier take the "potential landscape" approach, which is a natural bridge between probabilistic and dynamical perspectives.  If a Markov chain on a finite state space has a stationary distribution $\pi(\mathbf{x})$, one can define a potential (or energy) function[^15][^16]

$$
U(\mathbf{x}) = -\log \pi(\mathbf{x}),
$$

so that states with high probability have low energy (deep basins) and states with low probability have high energy (basin boundaries and hilltops).  The Ising model does this explicitly: the Boltzmann distribution $\pi(\mathbf{s}) \propto e^{-\beta H(\mathbf{s})}$ makes the energy function the Hamiltonian $H$, and the make_2d_Isingland function in your script computes exactly this potential aggregated over the number of active nodes.[^16][^15]

The VAR script takes the continuous analogue: the stationary distribution of a stable VAR($1$) process $\mathbf{x}_{t+1} = \boldsymbol{\mu} + B\mathbf{x}_t + \boldsymbol{\varepsilon}_t$ is multivariate normal with mean $(I-B)^{-1}\boldsymbol{\mu}$ and covariance $\Sigma_\infty$ satisfying the discrete Lyapunov equation $\Sigma_\infty = B\Sigma_\infty B^\top + \Sigma_\varepsilon$.  The potential over the sum score $S_t = \mathbf{1}^\top \mathbf{x}_t$ is then a Gaussian potential $U(s) = -\log \phi(s; \bar{s}, \sigma_S)$, where $\bar{s} = \mathbf{1}^\top(I-B)^{-1}\boldsymbol{\mu}$ and $\sigma_S^2 = \mathbf{1}^\top\Sigma_\infty \mathbf{1}$.[^16]

### 9.2 R Code: Unifying the Two Scripts — Shared Potential Landscape Framework

```{r}
# ============================================================
# Section 9: Unified potential landscape — Ising and VAR side by side
# ============================================================

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(123)

sigmoid <- function(z) 1 / (1 + exp(-z))

# ------------------------------------------------------------
# Part A: VAR potential landscape
# ------------------------------------------------------------
N <- 9

Beta <- matrix(sample(c(0, 1), N^2, replace = TRUE), N, N) *
  matrix(runif(N^2, 0.08, 0.22), N, N)
diag(Beta) <- runif(N, 0.2, 0.4)

Sigma_eps <- diag(N)

# Ensure stationarity before using the discrete Lyapunov solution
spectral_radius <- function(A) max(Mod(eigen(A, only.values = TRUE)$values))

if (spectral_radius(Beta) >= 0.98) {
  Beta <- Beta / (1.02 * spectral_radius(Beta))
}

lyap_solve <- function(B, S) {
  solve(diag(nrow(B)^2) - kronecker(B, B), as.vector(S))
}

Sigma_inf <- matrix(lyap_solve(Beta, Sigma_eps), N, N)

# Summary diagnostics for the baseline VAR
var_diag <- data.frame(
  spectral_radius = spectral_radius(Beta),
  trace_sigma = sum(diag(Sigma_inf)),
  det_sigma = det(Sigma_inf),
  min_eig_sigma = min(eigen(Sigma_inf, only.values = TRUE)$values),
  max_eig_sigma = max(eigen(Sigma_inf, only.values = TRUE)$values)
)

# Potential curves across coupling multipliers
mult_vals <- c(0.7, 1.0, 1.3)

var_potentials <- bind_rows(lapply(mult_vals, function(mult) {
  Bm <- Beta * mult
  rho <- spectral_radius(Bm)
  if (rho >= 0.98) {
    Bm <- Bm / (1.02 * rho)
  }
  
  Sig_m <- matrix(lyap_solve(Bm, Sigma_eps), N, N)
  mu_sum <- sum(rep(3, N))
  sd_sum <- sqrt(sum(Sig_m))
  x_vals <- seq(-5, 35, length.out = 500)
  U_vals <- -dnorm(x_vals, mean = mu_sum, sd = sd_sum, log = TRUE)
  
  data.frame(
    x = x_vals,
    U = U_vals,
    scenario = paste0("mult = ", mult),
    model = "VAR",
    x_min = mu_sum
  )
}))

var_summary <- var_potentials %>%
  group_by(scenario) %>%
  summarise(
    U_min = min(U),
    U_max = max(U),
    x_at_min = x[which.min(U)],
    .groups = "drop"
  )


# ------------------------------------------------------------
# Part B: Ising-type potential via mean-field approximation
# ------------------------------------------------------------
threshold_vals <- c(-1.0, 0.0, 1.0)
coupling_vals <- c(1.0, 2.0)

ising_grid <- expand.grid(
  threshold = threshold_vals,
  coupling = coupling_vals,
  stringsAsFactors = FALSE
)

ising_list <- vector("list", nrow(ising_grid))

for (i in seq_len(nrow(ising_grid))) {
  h <- ising_grid$threshold[i]
  w <- ising_grid$coupling[i]
  
  k_vals <- 0:9
  raw <- sapply(k_vals, function(k) {
    p <- sigmoid(-h + w * k / 9)
    dbinom(k, 9, p) + 1e-12
  })
  
  prob <- raw / sum(raw)
  
  ising_list[[i]] <- data.frame(
    x = k_vals,
    U = -log(prob),
    scenario = paste0("h = ", h, ", w = ", w),
    threshold = h,
    coupling = w,
    model = "Ising"
  )
}

ising_potentials <- bind_rows(ising_list)

ising_summary <- ising_potentials %>%
  group_by(scenario, threshold, coupling) %>%
  summarise(
    U_min = min(U),
    U_max = max(U),
    k_at_min = x[which.min(U)],
    entropy_like = -sum(exp(-U) * log(exp(-U))),
    .groups = "drop"
  )
# ------------------------------------------------------------
# Additional summary stats
# ------------------------------------------------------------
var_potential_stats <- var_potentials %>%
  group_by(scenario) %>%
  summarise(
    mean_U = mean(U),
    sd_U = sd(U),
    area_U = sum(U),
    .groups = "drop"
  )

ising_potential_stats <- ising_potentials %>%
  group_by(scenario) %>%
  summarise(
    mean_U = mean(U),
    sd_U = sd(U),
    area_U = sum(U),
    .groups = "drop"
  )

# ------------------------------------------------------------
# Plots
# ------------------------------------------------------------
p_var <- ggplot(var_potentials, aes(x = x, y = U, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(data = var_summary, aes(xintercept = x_at_min, colour = scenario),
             linetype = 2, alpha = 0.6, show.legend = FALSE) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank()) +
  labs(
    title = "VAR potential landscape",
    subtitle = "Lower curves indicate smaller negative log-density under the Gaussian sum-score approximation",
    x = "Sum score S",
    y = "U(S) = -log π(S)",
    colour = "Multiplier"
  )

p_ising <- ggplot(ising_potentials, aes(x = x, y = U, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.7) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank()) +
  labs(
    title = "Ising mean-field potential",
    subtitle = "Threshold and coupling shift the discrete energy profile over active-node counts",
    x = "Number of active nodes k",
    y = "U(k)",
    colour = "Setting"
  )

p_var_stats <- var_diag %>%
  mutate(metric = "spectral_radius") %>%
  ggplot(aes(x = metric, y = spectral_radius)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red4") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "VAR stability check",
    subtitle = "Spectral radius below 1 implies a stationary VAR(1) approximation",
    x = "",
    y = "Spectral radius"
  )

p_ising_stats <- ising_summary %>%
  ggplot(aes(x = scenario, y = k_at_min, fill = factor(coupling))) +
  geom_col(width = 0.7) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(
    title = "Ising modal active count",
    subtitle = "Location of the minimum depends on threshold and coupling",
    x = "Scenario",
    y = "k at minimum",
    fill = "Coupling"
  )

# Heatmap of VAR coefficients
Beta_df <- as.data.frame(as.table(Beta))
colnames(Beta_df) <- c("from", "to", "weight")

p_heat <- ggplot(Beta_df, aes(x = to, y = from, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  coord_equal() +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(
    title = "VAR coefficient matrix",
    subtitle = "Sparse random connectivity with diagonal self-coupling",
    x = "To node",
    y = "From node",
    fill = "Beta"
  )

# Combined figure
combined_plot <-
  (p_var | p_ising) /
  (p_var_stats | p_ising_stats) /
  p_heat +
  plot_layout(heights = c(1, 0.8, 1))

print(combined_plot)

# ------------------------------------------------------------
# Tables for reporting
# ------------------------------------------------------------
print(var_diag)
print(var_summary)
print(var_potential_stats)
print(ising_summary)
print(ising_potential_stats)

# ------------------------------------------------------------
# Optional exports
# ------------------------------------------------------------
# ggsave("unified_potential_landscape.png", combined_plot, width = 14, height = 16, dpi = 300)
# write.csv(var_summary, "var_potential_summary.csv", row.names = FALSE)
# write.csv(ising_summary, "ising_potential_summary.csv", row.names = FALSE)
```

This code places the Ising and VAR potential landscapes on a common conceptual footing.  In both cases, the shape of the potential — its number of minima, their depth, and their location — is entirely determined by the network parameters: weights and thresholds in the Ising case, and the coefficient matrix and innovation variance in the VAR case.  The basin of each minimum is the region of state space that flows downhill toward it, and transitions between basins require perturbations large enough to cross the energy barrier (the local maximum between minima).[^6][^15][^5][^16]

The code implements two complementary statistical frameworks for modeling psychopathological symptom networks: a continuous-time VAR(1) autoregressive system and a discrete Ising-type model under mean-field approximation. In the VAR formulation, symptom dynamics follow $\mathbf{x}_t = B\mathbf{x}_{t-1} + \boldsymbol{\varepsilon}_t$, where the coefficient matrix $B$ encodes directed pairwise influences between symptoms and $\boldsymbol{\varepsilon}_t$ represents independent Gaussian innovations . The asymptotic covariance $\Sigma_\infty$ is obtained via the discrete Lyapunov equation $\Sigma_\infty = B\Sigma_\infty B^\top + \Sigma_\varepsilon$, which admits a closed-form solution $\text{vec}(\Sigma_\infty) = (I_{N^2} - B \otimes B)^{-1}\text{vec}(\Sigma_\varepsilon)$ . This covariance structure determines the potential landscape $U(S) = -\log \mathcal{N}(S; \mu_S, \sigma_S^2)$ over the sum score $S = \sum_i x_i$, where shallower wells correspond to higher symptom variability and reduced network stability. The spectral radius constraint $\rho(B) < 1$ ensures stationarity, a prerequisite for interpreting $\Sigma_\infty$ as a steady-state distribution of symptom severity .

The Ising component adopts a binary-node representation where each symptom is either active or inactive, with activation probability governed by a local field $\phi_k = -h + w \cdot k/N$. Here, $h$ functions as an external threshold (analogous to diathesis or baseline vulnerability) and $w$ captures coupling strength between symptoms . Under the mean-field assumption, the probability of exactly $k$ active nodes is approximated as $P(k) \propto \binom{N}{k} p_k^k (1-p_k)^{N-k}$ with $p_k = \sigma(\phi_k)$, yielding a discrete potential $U(k) = -\log P(k)$. This formulation reveals how threshold-coupling interactions shape the location of the modal active count $k_{\text{min}}$: negative thresholds shift the minimum toward higher $k$ (promoting symptom co-activation), while stronger coupling sharpens the well and can induce bifurcation-like transitions between low- and high-symptom states . The entropy-like quantity $-\sum e^{-U} \log e^{-U}$ provides a measure of distributional dispersion, indexing the degree of symptom heterogeneity across the population.

Together, these models operationalize the potential landscape framework for psychopathology, where low-energy states correspond to stable symptom configurations and barrier heights reflect resistance to state transitions . The VAR approach captures continuous symptom severity and temporal autocorrelation, making it suitable for longitudinal data with graded measurements; the Ising model captures discrete activation patterns and threshold effects, aligning with categorical diagnostic criteria. By comparing potential curves across coupling multipliers (VAR) and threshold-coupling grids (Ising), the code enables systematic exploration of how network connectivity moderates symptom variability and stability. This dual perspective is particularly relevant for computational psychiatry, where symptom networks are hypothesized to exhibit critical dynamics near tipping points between health and disorder , and where intervention strategies might target high-influence nodes to reshape the global potential landscape.

## 10. Summary: The Full Causal Chain

The theoretical framework developed across these sections can be organized as a rigorous chain of four propositions.[^10][^1][^4]

**Proposition 1 (Structure → Transitions).** Network density $D$, edge-level persistence $b_1$, and transitivity $c$ enter the logistic transition model as additive predictors. Increasing density feedback ($a_1 > 0$) raises formation probabilities uniformly; increasing persistence ($b_1$) lowers dissolution probability; increasing transitivity ($c$) creates local reinforcement. These changes alter the full $N(N-1)/2$-dimensional transition kernel of the network Markov chain.[^4]

**Proposition 2 (Transitions → Markov Chain Geometry).** The transition kernel defines a Markov chain on the space of network configurations. Its stationary distribution $\pi$ concentrates on recurrent classes (attractors), and the basin structure is determined by the transition probabilities away from each region of state space. Density feedback can create bistability (two recurrent classes); persistence deepens individual basins; transitivity can create multiple community attractors.[^2][^1]

**Proposition 3 (Spectral Structure → Attractor Count).** The eigenvalues of the adjacency or coupling matrix $\mathbf{W}$ constrain which collective modes are self-reinforcing. Large positive eigenvalues correspond to cooperative modes that become attractors; a spectral gap between leading and trailing eigenvalues signals the number of dominant attractor basins. Network structural changes that alter the eigenspectrum (adding density, introducing block structure through transitivity) directly change the number and nature of attractors.[

**Proposition 4 (Continuous Analogy).** In continuous-time network-coupled ODEs, the Jacobian of the vector field at an equilibrium plays the role of the linearized transition matrix. Eigenvalues of the Jacobian with negative real parts confirm local basin structure; bifurcations occur when network parameters push eigenvalues across the imaginary axis, creating or destroying attractors. The AR coefficient in the discrete case and the coupling weights in the ODE case are both mechanisms that control this crossing.

The practical consequence for research and modeling is that one can read basin structure and transition dynamics off the network parameters without running full simulations: the spectral radius controls global stability, the sign of density dependence controls mono- versus bistability, the magnitude of persistence controls basin depth, and the block structure of the coupling matrix controls the number of competing attractors.  Your Ising and VAR scripts are complementary instantiations of this framework — one discrete with energy minima over configuration space, the other continuous with Gaussian potential over the sum score — both governed by the same underlying logic of network structure shaping long-run dynamical geometry.


[^1]: https://www.santafe.edu/research/results/working-papers/basins-of-attraction-in-network-dynamics-a-concept

[^2]: https://www.ddlab.com/downloads/papers/evodevo3.pdf

[^3]: https://users.sussex.ac.uk/~andywu/downloads/papers/2010_basins_of_attraction.pdf

[^4]: http://stats.lse.ac.uk/q.yao/qyao.links/paper/nwDE24.pdf

[^5]: https://www.sciencedirect.com/science/article/pii/S0022249625000690

[^6]: https://link.aps.org/doi/10.1103/PhysRevE.80.031110

[^7]: https://en.wikipedia.org/wiki/Bifurcation_theory

[^8]: http://www.professeurs.polymtl.ca/jerome.le-ny/teaching/CPS/Lec3/4_Stability_notes.pdf

[^9]: https://en.wikipedia.org/wiki/Lyapunov_stability

[^10]: http://www.scholarpedia.org/article/Hopfield_network

[^11]: https://pages.uoregon.edu/meravs/assets/files/Hopfield_Review.pdf

[^12]: https://link.aps.org/doi/10.1103/PhysRevE.108.014402

[^13]: https://link.aps.org/doi/10.1103/PhysRevResearch.5.033139

[^14]: https://pmc.ncbi.nlm.nih.gov/articles/PMC12559148/

[^15]: 250805_IsingPotentialsKendler-2.R

[^16]: 240916_VARpotentials.R

[^17]: https://direct.mit.edu/artl/article/18/3/325/2722/Exploring-Discrete-Dynamics-Andrew-Wuensche-2011

[^18]: https://www.alife.org/wp-content/uploads/2013/08/collections_ECAL93-1150-1159-Wuensche.pdf

[^19]: https://ideas.repec.org/p/wop/safiwp/98-11-101.html

[^20]: https://ideas.repec.org/a/eee/phsmap/v583y2021ics0378437121004817.html

[^21]: https://perso.uclouvain.be/raphael.jungers/sites/default/files/kcfinder/files/pathcomplete.pdf

[^22]: https://link.aps.org/doi/10.1103/PhysRevLett.98.198701

[^23]: https://yonsei.elsevierpure.com/en/publications/approximation-method-to-metastability-an-application-to-nonrevers

[^24]: https://arxiv.org/html/2510.21755v1

[^25]: https://en.wikipedia.org/wiki/Hopfield_network

[^26]: https://www.sciencedirect.com/topics/computer-science/hopfield-neural-networks

[^27]: https://ocw.mit.edu/courses/9-40-introduction-to-neural-computation-spring-2018/d0eb7c14d2fa8ddb6bed2583d2f2962d_MIT9_40S18_Lec20.pdf

[^28]: https://www.igi-global.com/article/optimization-of-hopfield-neural-network-for-improved-pattern-recall-and-storage-using-lyapunov-energy-function-and-hamming-distance/296592

[^29]: https://en.wikipedia.org/wiki/Modern_Hopfield_network

[^30]: https://arxiv.org/html/2605.03557v1

[^31]: https://arxiv.org/html/2505.01218v5

[^32]: https://www.cs.cmu.edu/afs/cs/academic/class/15782-f06/slides/hopfield.pdf

[^33]: https://www.math.uwaterloo.ca/~sacampbe/preprints/siam_shayer.pdf

[^34]: http://webhome.phy.duke.edu/~nb170/teaching/physics567/111-Hopfield.pdf



